---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r warning=FALSE, echo = TRUE, results = 'hide', message = FALSE}
library(MAST)
library(data.table)
library(R.utils)
library(doParallel)
library(stringr)
```

# Processing data

```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../../scRNAseq/Sade_Feldman_2018/", fnm1), 
                fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../../scRNAseq/Sade_Feldman_2018/", fnm2), 
                fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
rm(sf_tpm1)
rm(sf_tpm2)

sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
```

```{r}
dim(sf_tpm) # 55737 x 16292
sf_tpm[1:2,1:4]

gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

Only select classified cells  
```{r}
cell_type = read.table("./output/cell_type.txt", header = TRUE, 
                       sep = "\t", as.is = TRUE)
dim(cell_type)
cell_type[1:2,]

table(cell_type$type, useNA="ifany")
clustered_cells = cell_type$cell[cell_type$type != "unclustered"]
cluster_sf_tpm  = sf_tpm[, ..clustered_cells]
dim(cluster_sf_tpm)
rownames(cluster_sf_tpm) = gene.names
cluster_sf_tpm[1:3,1:3]

rm(sf_tpm)
```

## percentages of zero expression genes for each cell in each cell type
```{r}
ctype_clustered  = factor(cell_type$type[match(clustered_cells, cell_type$cell)])
zero.percentages = data.frame(percentage.of.zero = colMeans(cluster_sf_tpm == 0), 
                              grouping = ctype_clustered)
dim(zero.percentages)
head(zero.percentages)

fun1 <- function(x) { rownames(x)[order(x$percentage.of.zero)][1:200] }
fewzero_cells = unname(unlist(by(zero.percentages, zero.percentages$grouping, 
                                 FUN = fun1)))

cluster_sf_tpm = cluster_sf_tpm[, ..fewzero_cells]
cell_type = cell_type[match(fewzero_cells, cell_type$cell),]
dim(cell_type)
cell_type[1:2,]
```

# Select ~10,000 genes expressed in at least a proportion of cells

Calculate the proportion of cells that each gene is expressed in
```{r}
cellprop.per.gene = rowMeans(cluster_sf_tpm > 0)
summary(cellprop.per.gene)

he.genes = gene.names[cellprop.per.gene > 0.05] 
HEgenes_cluster_tpm = cluster_sf_tpm[cellprop.per.gene > 0.05, ]
rownames(HEgenes_cluster_tpm) = he.genes
dim(HEgenes_cluster_tpm)

summary(cellprop.per.gene[cellprop.per.gene > 0.05])

rm(cellprop.per.gene)
rm(cluster_sf_tpm)
```

# Average expression of each gene in each cell type

First calculate the expression of each gene across all cells

```{r}
all.expression = rowSums(HEgenes_cluster_tpm)
```

u_1 - u_11: Average expression of each gene for each cell type
o_1 - o_11: For a particular cell type, the average expression of each gene in all other cells

```{r}
average_expression = data.frame(gene = he.genes)
levels(factor(cell_type$type))

for (i in levels(zero.percentages$grouping)) {
  # select cells in each cell cluster
  cc = cell_type$cell[cell_type$type == i] 
  # total expression of each gene across all the cells of a particular cluster
  cluster.exp = apply(HEgenes_cluster_tpm[, ..cc], 1, sum)
  # total expression of each gene across all other cells
  exclu.exp = all.expression - cluster.exp
  # divide by the number of cells in the cluster to get the average expression
  aveg.cluster.exp = cluster.exp / length(cc)
  # divide by the number of all other cells to get the average expression
  aveg.exclu.exp = exclu.exp / (ncol(HEgenes_cluster_tpm) - length(cc))
  
  fc = aveg.exclu.exp / aveg.cluster.exp
  
  ulabel = paste0("u_",i)
  average_expression[[ulabel]] = aveg.cluster.exp
  
  olabel = paste0("o_",i)
  average_expression[[olabel]] = aveg.exclu.exp
  
  fclabel = paste0("fc_",i)
  average_expression[[fclabel]] = fc
}

rm(cc)
rm(cluster.exp)
rm(exclu.exp)
rm(aveg.cluster.exp)
rm(aveg.exclu.exp)
rm(fc)
for(i in 1:5){ gc() }
gc()
```

```{r}
write.table(average_expression, quote = FALSE, 
            file = "output/Average_GeneExpression_by_CellType_filtered_cells.txt", 
            sep = "\t", row.names = FALSE, col.names = TRUE)
```

# Testing differential gene expression

##  Loading and transforming data

```{r}
HEgenes_cluster_tpm_log = log2(1 + as.matrix(HEgenes_cluster_tpm))

row.names(HEgenes_cluster_tpm_log) = he.genes
cell.id = colnames(HEgenes_cluster_tpm_log)

dim(HEgenes_cluster_tpm_log)
HEgenes_cluster_tpm_log[1:2,1:5]
```

## Calculating the CDR rate: proportion of genes detected for each cell
```{r}
cdr = colMeans(HEgenes_cluster_tpm > 0)
summary(cdr)
```

## Preparing FromMatrix

```{r}
fData = data.frame(primerid = he.genes)
cData = data.frame(wellKey = cell.id)

sca = FromMatrix(HEgenes_cluster_tpm_log, cData, fData) 

colData(sca)$cngeneson = scale(cdr)

uniq_cell_types = sort(unique(cell_type$type))

for(i in 1:11){
  label = uniq_cell_types[i]
  colData(sca)[[label]] = "control"
  caseid = match(cell_type$cell[cell_type$type == label],colData(sca)$wellKey)
  colData(sca)[[label]][caseid] = "case"
  colData(sca)[[label]] = as.factor(colData(sca)[[label]])
}

colData(sca)

rm(HEgenes_cluster_tpm)
rm(HEgenes_cluster_tpm_log)
for(i in 1:5){ gc() }
gc()
```
  
## Fitting model

```{r}
# number of cores for multi-core computation
nCore = 10
registerDoParallel(cores=nCore)
options(mc.cores=nCore)
getOption("mc.cores")

mast_pval_zlm_df = fData
names(mast_pval_zlm_df)[1] = "gene"

for(i in 1:11){
  cti = uniq_cell_types[i]

  cat(i, cti, date(), "\n")
  f0  = as.formula(sprintf("~ %s + cngeneson", cti))
  b0  = zlm(formula = f0, sca = sca, parallel = TRUE) 
  cat("start lrtest", date(), "\n")

  lrt0 = lrTest(b0, cti)

  mast_pval_zlm_df[[cti]]  = apply(lrt0, 1, function(x){x[3,3]})
  gc()
}

dim(mast_pval_zlm_df)
mast_pval_zlm_df[1:2,]
write.csv(mast_pval_zlm_df, quote = FALSE, 
          file = "output/MastGenePvals_by_CellType_filtered_cells.csv")
```

```{r}
gc()
sessionInfo()
```




