---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r warning=FALSE, echo = TRUE, results = 'hide', message = FALSE}
library(lme4)
library(data.table)
library(R.utils)
library(stringr)
library(ggcorrplot)
```

# Check the DE resutls from MAST analysis

## Histogram of p-values

Histogram of p-values for each comparison
```{r fig.height=12, fig.width=8}
mast_pval_df = read.csv("./output/MastGenePvals_by_CellType.csv", 
                        header = TRUE, row.names = 1, as.is=TRUE)
head(mast_pval_df)

par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1), cex.main=0.8)
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))
for (i in 2:ncol(mast_pval_df)) {
  cti = names(mast_pval_df)[i]
  hist(mast_pval_df[,i], xlab = "p-values", breaks = 100, 
       main = paste("DE pvals of", cti, "vs. all other cells"))
}

par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1), cex.main=0.8)
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))
for (i in 2:ncol(mast_pval_df)) {
  cti = names(mast_pval_df)[i]
  pvi = mast_pval_df[,i]
  pvi[which(pvi < 1e-50)] = 1e-50
  hist(-log10(pvi), xlab = "p-values", breaks = 100, 
       main = paste("DE pvals of", cti, "vs. all other cells"))
}
```

## Volcano plot of p-values for each comparison

Read in dataframe of fold changes.
```{r}
fc_df = read.table("./output/Average_GeneExpression_by_CellType.txt", 
                   header = TRUE, sep = "\t", as.is=TRUE)
head(fc_df)
types = names(mast_pval_df)[-1]
types

table(fc_df$gene == mast_pval_df$gene)
fc_df = merge(fc_df, mast_pval_df, by="gene")
```

```{r fig.height=12, fig.width=8}
par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1))
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))

for (i in 1:length(types)) {
  cti  = types[i]
  pval = fc_df[[cti]]
  pval[which(pval < 1e-50)] = 1e-50
  log2_fc = log2(fc_df[[paste0("fc_",cti)]])
  log2_fc[which(log2_fc > 4)] = 4
  log2_fc[which(log2_fc < -4)] = -4
  
  plot(log2_fc, -log10(pval), 
       main = paste0("Volcano plot, ", types[i]), 
       xlab = "log2(fold change)", ylab = "-log10(p-value)", 
       xlim = c(-4,4), ylim = c(0, 50), cex=0.4, pch=20)
  abline(h=-log10(0.05/nrow(mast_pval_df)),col="red")
  abline(v=log2(1), col="red")
  abline(v=log2(1/2), col="red")
}
```

# Select differentially expressed genes for each cell type

Summarize the fold changes and mean expression for each gene. Find the minimum fc change for each gene. 

```{r}
fc_cols = grep("^fc_", names(fc_df), value=TRUE)
u_cols  = grep("^u_", names(fc_df), value=TRUE)

apply(fc_df[,fc_cols], 2, function(v){table(v < 0.5)})
apply(fc_df[,fc_cols], 2, function(v){table(v < 0.6)})

fc_mtrx = fc_df[,fc_cols]
min_fc  = apply(fc_mtrx, 1, min)
summary(min_fc)
hist(min_fc, xlab="minimum fold change per gene", main="", 
     sub="fc = mean(all other CTs) / mean(CT of interest)")

u_mtrx = fc_df[,u_cols]
u_mtrx[u_mtrx < 1e-3] = 1e-3
u_mtrx = log10(u_mtrx)
cr_mtrx = cor(u_mtrx, use="pair", method = "spearman")
summary(c(cr_mtrx))

ggcorrplot(cr_mtrx) + 
  scale_fill_gradient2(limit = c(0.6,1), low = "blue", high = "red", 
                       mid = "white", midpoint = 0.8)
```

Select around 60 genes per cell type. 

1. First, for each cell type, pre-filter genes with p-value passing the Bonferroni threshold, and fold change smaller than 0.5 or fold change < 0.75 while it is the minimum fold change across all cell types. 

2. Among those pre-filtered genes, select those genes with smallest p-value and fold change, by scanning a series of percentiles. 

Here fold changes is defined as the raio between the average expression from other cell types versus the average expression of the cell type of interest. For example, for 1 percentile, we select the intersection of the genes with p-value < 1 percentile and fold change smaller than 1 percentile. We choose  the percentile that gives us around 60 genes. 

```{r}
qs = seq(0.001, 1, by=0.001)

q2use = n2choose = rep(NA, 11)

for (i in 1:11) {
  cti = types[i]

  fc_i = fc_df[,paste0("fc_", cti)]
  pv_i = fc_df[[cti]]
  
  w2use_fc = fc_i < 0.5 | (fc_i == min_fc & fc_i < 0.75)
  w2use_v0 = which(pv_i < 0.05/nrow(mast_pval_df) & w2use_fc)
  
  fc_i = fc_i[w2use_v0]
  pv_i = pv_i[w2use_v0]
  
  ni = rep(NA, length(qs))
  
  for(j in 1:length(qs)){
    qj = qs[j]
    ni[j] = sum(fc_i <= quantile(fc_i, probs=qj) & 
                  pv_i <=  quantile(pv_i, probs=qj))
  }
  q2use[i]    = qs[which.min(abs(ni - 60))]
  n2choose[i] = ni[which.min(abs(ni - 60))]
}
data.frame(types, q2use, n2choose)
```

```{r}
genes.selected.perCellType = list()

for (i in 1:11) {
  cti = types[i]
  
  fc_i = fc_df[,paste0("fc_",cti)]
  pv_i = fc_df[[cti]]
  
  w2use_fc = fc_i < 0.5 | (fc_i == min_fc & fc_i < 0.75)
  w2use_v0 = which(pv_i < 0.05/nrow(mast_pval_df) & w2use_fc)
  
  fc_i = fc_i[w2use_v0]
  pv_i = pv_i[w2use_v0]

  fc_q = quantile(fc_i, probs=q2use[i])
  pv_q = quantile(pv_i, probs=q2use[i])
  
  w2use = w2use_v0[fc_i <= fc_q & pv_i <= pv_q]
  
  genes.selected.perCellType[[cti]] = fc_df$gene[w2use]
  marker = rep(0, nrow(fc_df))
  marker[w2use] = 1
  fc_df[[paste0("marker_",cti)]] = marker
}

genes.selected.perCellType

```

Check how many genes are selected by more than one cell type
```{r}
selected.genes = unlist(genes.selected.perCellType, use.names = FALSE) 
length(selected.genes)
t1 = table(selected.genes)
table(t1)
t1[t1 >= 3]
```

Save the output  
```{r}
sig.genes = sort(unique(selected.genes))
signature_gene_df = fc_df[match(sig.genes, fc_df$gene), 
                          c("gene", paste0("u_",types))]
colnames(signature_gene_df) = c("geneSymbol", types)
write.table(signature_gene_df, file = "./output/SF2018_sig_gene_matrix.txt", 
            quote = FALSE, sep = "\t", row.names = FALSE)

dim(fc_df)
fc_df[1:2,]

write.table(fc_df, file = "./output/SF2018_sig_gene_details.txt", 
            quote = FALSE, sep = "\t", row.names = FALSE)

```


```{r}
gc()
sessionInfo()
```
