---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r warning=FALSE, echo = TRUE, results = 'hide', message = FALSE}
library(lme4)
library(data.table)
library(R.utils)
library(stringr)
library(ggcorrplot)
```

# Check the DE resutls from MAST analysis

## Histogram of p-values

Histogram of p-values for each comparison
```{r fig.height=12, fig.width=8}
mast_pval_df = read.csv("./output/MastGenePvals_by_CellType.csv", 
                        header = TRUE, row.names = 1, as.is=TRUE)
head(mast_pval_df)

par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1), cex.main=0.8)
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))
for (i in 2:ncol(mast_pval_df)) {
  cti = names(mast_pval_df)[i]
  hist(mast_pval_df[,i], xlab = "p-values", breaks = 100, 
       main = paste("DE pvals of", cti, "vs. all other cells"))
}

par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1), cex.main=0.8)
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))
for (i in 2:ncol(mast_pval_df)) {
  cti = names(mast_pval_df)[i]
  pvi = mast_pval_df[,i]
  pvi[which(pvi < 1e-50)] = 1e-50
  hist(-log10(pvi), xlab = "p-values", breaks = 100, 
       main = paste("DE pvals of", cti, "vs. all other cells"))
}
```

## Volcano plot of p-values for each comparison

Read in dataframe of fold changes.
```{r}
fc_df = read.table("./output/Average_GeneExpression_by_CellType.txt", 
                   header = TRUE, sep = "\t", as.is=TRUE)
head(fc_df)
types = names(mast_pval_df)[-1]
types

table(fc_df$gene == mast_pval_df$gene)
fc_df = merge(fc_df, mast_pval_df, by="gene")
```

```{r fig.height=12, fig.width=8}
par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1))
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))

for (i in 1:length(types)) {
  cti = types[i]
  pval = fc_df[[cti]]
  pval[which(pval < 1e-100)] = 1e-100
  log2_fc = log2(fc_df[[paste0("fc_",cti)]])
  log2_fc[which(log2_fc > 6)] = 6
  log2_fc[which(log2_fc < -6)] = -6
  
  plot(log2_fc, -log10(pval), 
       main = paste0("Volcano plot, ", types[i]), 
       xlab = "log2(fold change)", ylab = "-log10(p-value)", 
       xlim = c(-6,6), ylim = c(0, 100), cex=0.4, pch=20)
  abline(h=-log10(0.05/nrow(mast_pval_df)),col="red")
  abline(v=log2(1),col="red")
}

```

# Select differentially expressed genes for each cell type

Select around 60 genes per cell type. Select those genes with smallest p-value and fold change, by scanning a series of percentiles. We selected those with smallest fold changes because we define fold changes as the average expression from other cell types versus the average expression of the cell type of interest. For example, for 1 percentile, we select the intersection of the genes with p-value < 1 percentile and fold change smaller than 1 percentile. Choose the percentile that gives use between 55 and 65 genes. In addition, also require the fold change to be smaller than 0.75

```{r}

fc_cols = grep("^fc_", names(fc_df), value=TRUE)
u_cols  = grep("^u_", names(fc_df), value=TRUE)

apply(fc_df[,fc_cols], 2, function(v){table(v < 0.5)})
apply(fc_df[,fc_cols], 2, function(v){table(v < 0.7)})

u_mtrx = fc_df[,u_cols]
u_mtrx[u_mtrx < 1e-3] = 1e-3
u_mtrx = log10(u_mtrx)
cr_mtrx = cor(u_mtrx, use="pair", method = "spearman")
summary(c(cr_mtrx))
ggcorrplot(cr_mtrx) + 
  scale_fill_gradient2(limit = c(0.6,1), low = "blue", high =  "red", mid = "white", midpoint = 0.8)

qs = seq(0.001, 0.05, by=0.001)

q2use = n2choose = rep(NA, 11)

for (i in 1:11) {
  fc_i = fc_df[,paste0("fc_",types[i])]
  pv_i = fc_df[[types[i]]]
  
  ni = rep(NA, length(qs))
  
  for(j in 1:length(qs)){
    qj = qs[j]
    ni[j] = sum(fc_i <= quantile(fc_i, probs=qj) & 
                  pv_i <=  quantile(pv_i, probs=2*qj))
  }
  q2use[i]    = qs[which.min(abs(ni- 60))]
  n2choose[i] = ni[which.min(abs(ni- 60))]
}
data.frame(types, q2use, n2choose)
```

```{r}
genes.selected.perCellType = list()

for (i in 1:11) {
  type_i = types[i]
  fc_i = fc_df[,paste0("fc_",types[i])]
  fc_q = quantile(fc_i ,probs=q2use[i])
  pv_i = fc_df[[type_i]]
  pv_q = quantile(pv_i, probs=q2use[i])
  
  w2use = fc_i <= fc_q & pv_i <= pv_q
  genes.selected.perCellType[[type_i]] = fc_df$gene[w2use]
  marker = rep(0, nrow(fc_df))
  marker[w2use] = 1
  fc_df[[paste0("marker_",types[i])]] = marker
}

genes.selected.perCellType

```

Check how many genes are selected by more than one cell type
```{r}
selected.genes = unlist(genes.selected.perCellType, use.names = FALSE) 
length(selected.genes)
t1 = table(selected.genes)
table(t1)
t1[t1 >= 3]
```

Save the output  
```{r}
sig.genes = sort(unique(selected.genes))
signature_gene_df = fc_df[match(sig.genes, fc_df$gene), 
                          c("gene", paste0("u_",types))]
colnames(signature_gene_df) = c("geneSymbol", types)
write.table(signature_gene_df, file = "./output/SF2018_sig_gene_matrix.txt", 
            quote = FALSE, sep = "\t", row.names = FALSE)

dim(fc_df)
fc_df[1:2,]

write.table(fc_df, file = "./output/SF2018_sig_gene_details.txt", 
            quote = FALSE, sep = "\t", row.names = FALSE)

```


```{r}
gc()
sessionInfo()
```
