---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r warning=FALSE, echo = TRUE, results = 'hide', message = FALSE}
library(lme4)
library(data.table)
library(R.utils)
library(stringr)
```

# Check the DE resutls from MAST analysis

## Histogram of p-values

Histogram of p-values for each comparison
```{r fig.height=12, fig.width=8}
mast_pval_df = read.csv("./output/MastGenePvals_by_CellType_filtered_cells.csv", 
                        header = TRUE, row.names = 1, as.is=TRUE)
head(mast_pval_df)

par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1), cex.main=0.8)
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))
for (i in 2:ncol(mast_pval_df)) {
  cti = names(mast_pval_df)[i]
  hist(mast_pval_df[,i], xlab = "p-values", breaks = 500, 
       main = paste("DE pvals of", cti, "vs. all other cells"))
}
```

## Volcano plot of p-values for each comparison

Read in dataframe of fold changes.
```{r}
fc_df = read.table("./output/Average_GeneExpression_by_CellType_filtered_cells.txt", 
                   header = TRUE, sep = "\t", as.is=TRUE)
head(fc_df)
types = names(mast_pval_df)[-1]
types

table(fc_df$gene == mast_pval_df$gene)
fc_df = merge(fc_df, mast_pval_df, by="gene")
```

```{r fig.height=12, fig.width=8}
par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1))
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))

for (i in 1:length(types)) {
  cti = types[i]
  pval = fc_df[[cti]]
  pval[which(pval < 1e-100)] = 1e-100
  log2_fc = log2(fc_df[[paste0("fc_",cti)]])
  log2_fc[which(log2_fc > 6)] = 6
  log2_fc[which(log2_fc < -6)] = -6
  
  plot(log2_fc, -log10(pval), 
       main = paste0("Volcano plot, ", types[i]), 
       xlab = "log2(fold change)", ylab = "-log10(p-value)", 
       xlim = c(-6,6), ylim = c(0, 100), cex=0.4, pch=20)
  abline(h=-log10(0.05/nrow(mast_pval_df)),col="red")
  abline(v=log2(1),col="red")
}

```

# Select differentially expressed genes for each cell type

Select around 60 genes per cell type. Select those genes with smallest p-value and fold change, by scanning a series of percentiles. For example, for 1 percentile, we select the intersection of the genes with p-value < 1 percentile and fold change smaller than 1 percentile. Choose the percentile that gives use between 55 and 65 genes. 

```{r}
qs = seq(0.001, 0.05, by=0.001)

q2use = n2choose = rep(NA, 11)

for (i in 1:11) {
  fc_i = fc_df[,paste0("fc_",types[i])]
  pv_i = fc_df[[types[i]]]
  
  ni = rep(NA, length(qs))
  
  for(j in 1:length(qs)){
    qj = qs[j]
    ni[j] = sum(fc_i <= quantile(fc_i, probs=qj) & 
                  pv_i <=  quantile(pv_i, probs=qj))
  }
  q2use[i]    = qs[which.min(abs(ni- 60))]
  n2choose[i] = ni[which.min(abs(ni- 60))]
}
data.frame(types, q2use, n2choose)
```

```{r}
genes.selected.perCellType = list()

for (i in 1:11) {
  type_i = types[i]
  fc_i = fc_df[,paste0("fc_",types[i])]
  fc_q = quantile(fc_i ,probs=q2use[i])
  pv_i = fc_df[[type_i]]
  pv_q = quantile(pv_i, probs=q2use[i])
  w2use = fc_i <= fc_q & pv_i <= pv_q
  genes.selected.perCellType[[type_i]] = fc_df$gene[w2use]
  marker = rep(0, nrow(fc_df))
  marker[w2use] = 1
  fc_df[[paste0("marker_",types[i])]] = marker
}

genes.selected.perCellType
```

Check how many genes are selected by more than one cell type
```{r}
selected.genes = unlist(genes.selected.perCellType, use.names = FALSE) 
length(selected.genes)
t1 = table(selected.genes)
table(t1)
t1[t1 >= 3]
```

Save the output  
```{r}
sig.genes = sort(unique(selected.genes))
signature_gene_df = fc_df[match(sig.genes, fc_df$gene), c("gene",paste0("u_",types))]
colnames(signature_gene_df) = c("geneSymbol", types)
write.table(signature_gene_df, 
            file = "./output/SF2018_sig_gene_matrix_filtered_cells.txt", 
            quote = FALSE, sep = "\t", row.names = FALSE)

dim(fc_df)
fc_df[1:2,]

write.table(fc_df, file = "./output/SF2018_sig_gene_details_filtered_cells.txt", 
            quote = FALSE, sep = "\t", row.names = FALSE)
```


```{r}
gc()
sessionInfo()
```
