---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r warning=FALSE, echo = TRUE, results = 'hide', message = FALSE}
library(data.table)
library(R.utils)
library(stringr)
library(ggcorrplot)
theme_set(theme_bw())
```

# Processing data

```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../../scRNAseq/Sade_Feldman_2018/", fnm1), 
                fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../../scRNAseq/Sade_Feldman_2018/", fnm2), 
                fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
rm(sf_tpm1)
rm(sf_tpm2)

sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
```

```{r}
dim(sf_tpm) # 55737 x 16292
sf_tpm[1:2,1:4]

gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

Only select classified cells. 
```{r}
cell_type = read.table("./output/cell_type.txt", header = TRUE, 
                       sep = "\t", as.is = TRUE)
dim(cell_type)
cell_type[1:2,]

table(cell_type$type, useNA="ifany")
clustered_cells = cell_type$cell[cell_type$type != "unclustered"]
cluster_sf_tpm  = sf_tpm[, ..clustered_cells]
rownames(cluster_sf_tpm) = gene.names

dim(cluster_sf_tpm)
cluster_sf_tpm[1:3,1:3]

rm(sf_tpm)
rm(clustered_cells)
```

# Select ~10,000 genes expressed in at least a proportion of cells

Calculate the proportion of cells that each gene is expressed in
```{r}
cellprop.per.gene = rowMeans(cluster_sf_tpm > 0)

he.genes = gene.names[cellprop.per.gene > 0.05]
HEgenes_cluster_tpm = cluster_sf_tpm[cellprop.per.gene > 0.05, ]
rownames(HEgenes_cluster_tpm) = he.genes
dim(HEgenes_cluster_tpm)

rm(cluster_sf_tpm)
rm(cellprop.per.gene)
```

# Read in the two versions of signature genes
```{r}
sig_gene = fread("./output/SF2018_sig_gene_matrix.txt")
sig_gene_filtered = fread("./output/SF2018_sig_gene_matrix_filtered_cells.txt")

dim(sig_gene)
dim(sig_gene_filtered)

sig_gene[1:2,]
sig_gene_filtered[1:2,]
```

# Check the correlations between these two versions of signature matrix. 

The signature of CD8T_B does change a lot after selecting the top 200 cells with highiest expression. In fact, CD8T_B using all cells has highest correation with NK3 using filered cells, and CD8T_B using filered cells has highest correlation with NK1 using all the cells. 

```{r}
genes_shared = intersect(sig_gene$geneSymbol, 
                         sig_gene_filtered$geneSymbol)

mat1 = match(genes_shared, sig_gene$geneSymbol)
mat2 = match(genes_shared, sig_gene_filtered$geneSymbol)

sig_mtrx = data.matrix(sig_gene[mat1, -1])
sig_mtrx_filtered = data.matrix(sig_gene_filtered[mat2, -1])

dim(sig_mtrx)
dim(sig_mtrx_filtered)

sig_mtrx[1:2,]
sig_mtrx_filtered[1:2,]

cr1 = cor(sig_mtrx, sig_mtrx_filtered)
summary(diag(cr1))
sort(diag(cr1))[1:5]
summary(c(cr1))
ggcorrplot(cr1, tl.cex = 6)

round(cr1[c(2:4,7:9),c(2:4,7:9)], 2)
cor(sig_mtrx[,"CD8T_B"], sig_mtrx_filtered[,"NK3"])
```

# Read in the details of DE analysis
```{r}
sig_info = fread("./output/SF2018_sig_gene_details.txt")
sig_info_filtered = fread("./output/SF2018_sig_gene_details_filtered_cells.txt")

dim(sig_info)
dim(sig_info_filtered)

sig_info[1:2,c(1:4,(ncol(sig_info)-3):ncol(sig_info)), with=F]
sig_info_filtered[1:2,c(1:4,(ncol(sig_info)-3):ncol(sig_info)), with=F]
```

# illustrate the results for a few fenes
```{r, fig.asp = 0.5}
cluster_ct = cell_type$type[match(names(HEgenes_cluster_tpm), cell_type$cell)]
marker_col = grep("^marker_", names(sig_info), value = TRUE)

plot1 <- function(gene1, edata, cluster_ct, sig_info){
  y      = as.numeric(edata[which(rownames(edata) == gene1),])
  df1    = data.frame(expression = y, cell_type = cluster_ct)
  
  g1 = ggplot(df1, aes(x=cell_type, y=expression, fill=cell_type)) +
    geom_boxplot() + coord_flip() + theme(legend.position = "none") +
    ggtitle(gene1)
  print(g1)
  
  ct = marker_col[unlist(sig_info[gene==gene1, marker_col, with=FALSE])==1]
  print(ct)
}

plot1("FOXP3", HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("GZMA",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("PDCD1", HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("CTLA4", HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("LAG3",  HEgenes_cluster_tpm, cluster_ct, sig_info)

plot1("CD3E",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("CD4",   HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("CD8A",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("CD33",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("CD14",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("CD19",  HEgenes_cluster_tpm, cluster_ct, sig_info)

plot1("LMNA",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("FASLG",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("VCAM1",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("LEF1",  HEgenes_cluster_tpm, cluster_ct, sig_info)

setdiff(sig_info$gene[sig_info$marker_NK1 == 1], 
        sig_info_filtered$gene[sig_info_filtered$marker_NK1 == 1])

intersect(sig_info$gene[sig_info$marker_NK1 == 1], 
          sig_info_filtered$gene[sig_info_filtered$marker_NK1 == 1])

plot1("CCNB2", HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("CCNA2", HEgenes_cluster_tpm, cluster_ct, sig_info)

setdiff(sig_info$gene[sig_info$marker_NK2 == 1], 
        sig_info_filtered$gene[sig_info_filtered$marker_NK2 == 1])

intersect(sig_info$gene[sig_info$marker_NK2 == 1], 
          sig_info_filtered$gene[sig_info_filtered$marker_NK2 == 1])

plot1("NFKB1",  HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("FCGR3A", HEgenes_cluster_tpm, cluster_ct, sig_info)

setdiff(sig_info$gene[sig_info$marker_NK3 == 1], 
        sig_info_filtered$gene[sig_info_filtered$marker_NK3 == 1])

intersect(sig_info$gene[sig_info$marker_NK3 == 1], 
          sig_info_filtered$gene[sig_info_filtered$marker_NK3 == 1])

plot1("CCL4L2", HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("CCL4",   HEgenes_cluster_tpm, cluster_ct, sig_info)
plot1("GZMK",   HEgenes_cluster_tpm, cluster_ct, sig_info)

```


```{r}
gc()
sessionInfo()
```
