---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r eval=FALSE}
library("MAST")
library(data.table)
library(R.utils)  
```

# Processing data

```{r eval=FALSE}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
rm(sf_tpm1)
rm(sf_tpm2)

sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)

```

```{r eval=FALSE}
#dim(sf_tpm) # 55737 x 16292
#cell.names = colnames(sf_tpm)[-1]
gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

Only select classified cells  
```{r eval=FALSE}
cell_type = read.table("cell_type.txt",header = TRUE,sep = "\t",as.is=TRUE)
dim(cell_type)
cell_type[1:2,]
table(cell_type$type)
clustered_cells = cell_type$cell[cell_type$type != "unclustered"]
cluster_sf_tpm = sf_tpm[, ..clustered_cells] # 55737 x 14777
rownames(cluster_sf_tpm) = gene.names
cluster_sf_tpm[1:5,1:5]
rm(sf_tpm)
```

## percentages of zero expression genes for each cell in each cell type
```{r eval=FALSE}
zeros.per.cell = colSums(cluster_sf_tpm==0) / nrow(cluster_sf_tpm)
zero.percentages = data.frame(percentage.of.zero = zeros.per.cell, grouping = factor(cell_type$type[match(clustered_cells, cell_type$cell)]))
head(zero.percentages)
rm(zeros.per.cell)
fewzero_cells = unname(unlist(by(zero.percentages, zero.percentages$grouping, FUN = function(x) rownames(x)[order(x$percentage.of.zero)][1:100])))

cluster_sf_tpm = cluster_sf_tpm[, ..fewzero_cells]
```

# Select ~10,000 genes expressed in at least a proportion of cells

Calculate the proportion of cells that each gene is expressed in
```{r eval=FALSE}
cells.per.gene = rowSums(cluster_sf_tpm > 0)
cellprop.per.gene = cells.per.gene/ncol(cluster_sf_tpm)
summary(cellprop.per.gene)

he.genes = gene.names[cellprop.per.gene > 0.02] # kept 17066 genes
HEgenes_cluster_tpm = cluster_sf_tpm[cellprop.per.gene > 0.02, ] # 17066 x 1100
rownames(HEgenes_cluster_tpm) = he.genes

rm(cellprop.per.gene)
rm(cells.per.gene)
rm(cluster_sf_tpm)

cells.per.gene = rowSums(HEgenes_cluster_tpm > 0)
cellprop.per.gene = cells.per.gene/ncol(HEgenes_cluster_tpm)
summary(cellprop.per.gene)
table(cellprop.per.gene > 0.5)
```

# Average expression of each gene in each cell type

First calculate the expression of each gene across all cells

```{r eval=FALSE}
all.expression = apply(HEgenes_cluster_tpm, 1, sum)
```

u_1 - u_11: Average expression of each gene for each cell type
o_1 - o_11: For a particular cell type, the average expression of each gene in all other cells

```{r eval=FALSE}
average_expression = data.frame(gene = he.genes)

# order of 1-11: 
#   [1] "B cells"               "CD4 T.mem cells"       "CD4 T.reg cells"      
#  [4] "CD8_B cells"           "CD8_G cells"           "Dendritic cells"      
#  [7] "Monocytes/Macrophages" "NK1 cells"             "NK2 cells"            
# [10] "NK3 cells"             "Plasma cells"  

for (i in levels(zero.percentages$grouping)) {
  cc = fewzero_cells[cell_type$type[match(fewzero_cells, cell_type$cell)] == i] # select cells in each cell cluster
  cluster.exp = apply(HEgenes_cluster_tpm[, ..cc], 1, sum) # total expression of each gene across all cells in a particular cluster
  exclu.exp = all.expression - cluster.exp # total expression of each gene across all other cells
  aveg.cluster.exp = cluster.exp / length(cc) # divide by the number of cells in the cluster to get the average expression
  aveg.exclu.exp = exclu.exp / (ncol(HEgenes_cluster_tpm) - length(cc)) # divide by the number of all other cells to get the average expression
  fc = aveg.exclu.exp / aveg.cluster.exp
  ulabel = paste0("u_",i)
  average_expression[[ulabel]] = aveg.cluster.exp
  olabel = paste0("o_",i)
  average_expression[[olabel]] = aveg.exclu.exp
  fclabel = paste0("fc_",i)
  average_expression[[fclabel]] = fc
}

rm(cc)
rm(cluster.exp)
rm(exclu.exp)
rm(aveg.cluster.exp)
rm(aveg.exclu.exp)
rm(fc)
```

```{r eval=FALSE}
write.table(average_expression, file = "Average_GeneExpression_by_CellType_few0.txt", quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
```

# Testing differential gene expression

##  Loading and transforming data

```{r eval=FALSE}
# cluster_sf_tpm is a big data table of TPM data, with one row per gene and one column per cell
HEgenes_cluster_tpm_log = log2(1 + as.matrix(HEgenes_cluster_tpm)) #log transformed data

dim(HEgenes_cluster_tpm_log) # 17066 x 1000
HEgenes_cluster_tpm_log[1:2,1:5]

row.names(HEgenes_cluster_tpm_log) = he.genes

# get the cell names from the data
cell.id = colnames(HEgenes_cluster_tpm_log)   
# gene names are saved in the vector gene.names
```

## Calculating the CDR rate: proportion of genes detected for each cell
```{r eval=FALSE}
cdr = colMeans(HEgenes_cluster_tpm > 0)
summary(cdr)
```

## Preparing FromMatrix

```{r eval=FALSE}
fData = data.frame(primerid = he.genes)
cData = data.frame(wellKey = cell.id)
#nrow(HEgenes_cluster_tpm_log) == nrow(fData)
#ncol(HEgenes_cluster_tpm_log) == nrow(cData)

sca = FromMatrix(HEgenes_cluster_tpm_log, cData, fData) 
# On bayes.biostat.washington.edu:
# class: SingleCellAssay 
# dim: 17066 1100 
# metadata(0):
# assays(1): et
# rownames(17066): DPM1 SCYL3 ... RP11-225H22.7 RP11-180C16.1
# rowData names(1): primerid
# colnames(1100): E4_P4_M47 E5_P4_M47 ... G8_P1_M50_L001 H12_P2_M51_L001
# colData names(1): wellKey
# reducedDimNames(0):
# spikeNames(0):

colData(sca)$cngeneson = scale(cdr)
for(i in 1:11){
  label = paste0("G_",i)
  colData(sca)[[label]] = "control"
  caseid = match(fewzero_cells[cell_type$type[match(fewzero_cells, cell_type$cell)] == levels(zero.percentages$grouping)[i]],
                 colData(sca)$wellKey)
  colData(sca)[[label]][caseid] = "case"
  colData(sca)[[label]] = as.factor(colData(sca)[[label]])
}

colData(sca)
```
  
## Fitting model

```{r eval=FALSE}
zlmlist = list()
date() # "Mon Oct 26 01:36:48 2020"
zlmlist[["G_1"]] = zlm(formula = ~ G_1 + cngeneson, sca = sca, parallel = FALSE) 
date() # "Fri Oct 9 14:28:58 2020"
zlmlist[["G_2"]] = zlm(formula = ~ G_2 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_3"]] = zlm(formula = ~ G_3 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_4"]] = zlm(formula = ~ G_4 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_5"]] = zlm(formula = ~ G_5 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_6"]] = zlm(formula = ~ G_6 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_7"]] = zlm(formula = ~ G_7 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_8"]] = zlm(formula = ~ G_8 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_9"]] = zlm(formula = ~ G_9 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_10"]] = zlm(formula = ~ G_10 + cngeneson, sca = sca, parallel = FALSE)
zlmlist[["G_11"]] = zlm(formula = ~ G_11 + cngeneson, sca = sca, parallel = FALSE)
date() # "Mon Oct 26 02:07:38 2020"
zlmlist
```
     
## Testing

```{r eval=FALSE}
date() # "Mon Oct 26 02:28:50 2020"
lrtlist = list()
for (i in 1:11) {
  lrtlist[[paste0("G_",i)]] = lrTest(zlmlist[[i]], paste0("G_",i))
}
date() # "Mon Oct 26 02:54:42 2020"
str(lrtlist)
```
     
## Get the hurdel p-values

```{r eval=FALSE}
dim(lrtlist[[1]]) 
lrtlist[[1]][1, , ]
#  metric
# test.type     lambda df Pr(>Chisq)
#    cont   4.60528392  1 0.03187357
#    disc   0.07389619  1 0.78574618
#    hurdle 4.67918011  2 0.09636714
lrtlist[[1]][1, , ][3,3] # 0.09636714

mast_pval_glm_list = lapply(lrtlist, FUN = function(lrt0) apply(lrt0, 1, function(x){x[3,3]}))
str(mast_pval_glm_list)

mast_pval_glm_df = as.data.frame(mast_pval_glm_list)
write.csv(mast_pval_glm_df, file = "MastGenePvals_by_CellType_few0.csv", quote = FALSE)
```

## Histogram of p-values

Dataframe of p-values for each comparison
```{r}
mast_pval_df = read.csv("./output/MastGenePvals_by_CellType_few0.csv", header = TRUE, row.names = 1)
head(mast_pval_df)
```

Histogram of p-values for each comparison
```{r fig.height=12, fig.width=8}
typeslabel = c("B cells","CD4 T.mem cells","CD4 T.reg cells","CD8_B cells","CD8_G cells","Dendritic cells", "Monocytes/Macrophages", "NK1 cells", "NK2 cells", "NK3 cells", "Plasma cells")

par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1), cex.main=0.8)
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))
for (i in 1:11) {
  hist(mast_pval_df[,i], main = paste0("Diff. gene expression pvals when comparing ", typeslabel[i], " and all other cells"), xlab = "p-values", breaks = 500)
}
```

## Volcano plot of p-values for each comparison

Dataframe of fold changes
```{r}
fc_df = read.table("./output/Average_GeneExpression_by_CellType_few0.txt", header = TRUE, sep = "\t")
head(fc_df)
# extract column names
types = unique(sapply(strsplit(colnames(fc_df)[-1], c("[fcou]+_")),`[[`,2))
types
```

```{r fig.height=12, fig.width=8}
par(mgp=c(2,0.5,0), mar=c(3.5,3.5,1.5,1))
layout(matrix(1:12,nrow=6,ncol = 2,byrow = TRUE))
for (i in 1:11) {
  plot(log2(fc_df[,paste0("fc_",types[i])]), -log10(mast_pval_df[,i]), main = paste0("Volcano plot, ", types[i]), xlab = "log2(fold change)", ylab = "-log10(p-value)", xlim=c(-10,10), pch=20)
  abline(h=-log10(2.929802e-06),col="red")
  abline(v=log2(1),col="red")
}

```

# Select differentially expressed genes for each cell type

Select around 60 genes per cell type. Select those genes with smallest p-value and fold change, by scanning a series of percentiles. For example, for 1 percentile, we select the intersection of the genes with p-value < 1 percentile and fold change smaller than 1 percentile. Choose the percentile that gives use between 55 and 65 genes. 

```{r}
#for (i in 1:11) {
#  print(sum((fc_df[,paste0("fc_",types[i])] <= 
               quantile(fc_df[,paste0("fc_",types[i])],probs=seq(0,0.1,0.01))[5] 
             & mast_pval_df[,i] <= 
               quantile(mast_pval_df[,i],probs=seq(0,0.1,0.01))[5])))
#}

#j=5
#length(fc_df$gene[fc_df[,paste0("fc_",types[j])] <= quantile(fc_df[,paste0("fc_",types[j])],probs=0.042) & mast_pval_df[,j] <= quantile(mast_pval_df[,j],probs=0.042)])
```

```{r}
genes.selected.perCellType = list()

quantiles = c(0.011, 0.044, 0.018, 0.014, 0.042, 0.009, 0.018, 0.015, 0.027, 0.02, 0.027)

for (i in 1:11) {
  genes.selected.perCellType[[colnames(mast_pval_df)[i]]] = as.character(fc_df$gene[fc_df[,paste0("fc_",types[i])] <= quantile(fc_df[,paste0("fc_",types[i])],probs=quantiles[i]) & mast_pval_df[,i] <= quantile(mast_pval_df[,i],probs=quantiles[i])])
}

genes.selected.perCellType
```

Check how many genes are selected by more than one cell type
```{r}
selected.genes = unlist(genes.selected.perCellType, use.names = FALSE) # 701 genes in total
selected.genes[duplicated(selected.genes)] # 80 genes are selected in more than 1 cell type
```

Save the output  
```{r eval=FALSE}
sig.genes = sort(unique(selected.genes)) # 621 signature genes selected
signature_gene_df = fc_df[match(sig.genes,as.character(fc_df$gene)),c("gene",paste0("u_",types))]
colnames(signature_gene_df) = c("geneSymbol", types)
write.table(signature_gene_df, file = "./output/SF2018_few0_sig_gene_matrix.txt", quote = FALSE, sep = "\t", row.names = FALSE)
```

```{r}
gc()
sessionInfo()
```




