---
title: "Analysis of neoantigen data"
output:
  html_document:
    fig_width: 4
    fig_height: 3
    df_print: paged
    keep_md: true
---

# Read in PEPPRMINT estimates
```{r warning=FALSE, echo = FALSE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(ggplot2)
library(ggpointdensity)
library(readxl)
library(survival)
library(survminer)
library(dplyr)
library(viridis)

theme_set(theme_classic())
```

# Read in neoantigen estimation by NetMHC
```{r}
netmhc_all  = readRDS("../data/neoAg_netMHCpan4_1.rds")

colnames(netmhc_all)
dim(netmhc_all)
head(netmhc_all)
netmhc_all  = as.data.table(netmhc_all)
```

## Taking maximum for each somatic mutation using mutated petides
```{r}
netmhc_mut = netmhc_all[,.(EL_score_mut_max = max(EL_score_mut)), 
                        by = .(ID, sample)]
dim(netmhc_mut)
netmhc_mut
netmhc_mut$key = gsub("_", ":", netmhc_mut$ID)
netmhc_mut$sample = gsub("_hlai", "", netmhc_mut$sample)

g1 = ggplot(netmhc_mut, aes(x=EL_score_mut_max)) + xlab("NetMHC score (mut)") + 
  geom_histogram(color="darkblue", fill="lightblue")
g1
```

## Taking maximum for each somatic mutation using reference peptides
```{r}
netmhc_ref = netmhc_all[,.(EL_score_ref_max = max(EL_score_ref)), 
                        by = .(ID, sample)]
dim(netmhc_ref)
netmhc_ref
netmhc_ref$key = gsub("_", ":", netmhc_ref$ID)
netmhc_ref$sample = gsub("_hlai", "", netmhc_ref$sample)

g1 = ggplot(netmhc_ref, aes(x=EL_score_ref_max)) + xlab("NetMHC score (ref)") + 
  geom_histogram(color="darkblue", fill="lightblue")
g1
```

## Compare NetMHC estimates between reference and mutation
```{r fig.height=3.5, fig.width=4.5}
table(netmhc_mut$ID == netmhc_ref$ID)
table(netmhc_mut$sample == netmhc_ref$sample)

netmhc = merge(netmhc_mut, netmhc_ref)
netmhc

gg1 = ggplot(data = netmhc, mapping = aes(x = EL_score_ref_max, 
                                          y = EL_score_mut_max)) +
  geom_pointdensity(size=1) + xlab("NetMHCpan-4.1 ref") + 
  ylab("NetMHCpan-4.1 mut") + 
  scale_color_viridis() + theme(legend.position = "top")

pdf("step4_nb_analysis_files/netmhc_ref_vs_mut.pdf", width=3.5, height=4.2)
gg1
dev.off()

```

# Read in neoantigen estimation by PEPPRMINT

## Results after taking maximum for the scores from mutated peptides
```{r}
ppmint_mut = fread("../output_PEPPRMINT/PEPPRMINT_Riaz_2017_aggregate.tsv")
dim(ppmint_mut)
ppmint_mut
summary(ppmint_mut$pepprmint)

setequal(ppmint_mut$sample, netmhc$sample)
length(setdiff(ppmint_mut$key, netmhc$key))

g1 = ggplot(ppmint_mut, aes(x=pepprmint)) + xlab("PEPPRMINT score (mut)") + 
  geom_histogram(color="darkblue", fill="lightblue")
g1
```

## Results after taking maximum for the scores from reference peptides
```{r}
ppmint_ref = fread("../output_PEPPRMINT/PEPPRMINT_Riaz_2017_ref_aggregate.tsv")
dim(ppmint_ref)
ppmint_ref
summary(ppmint_ref$pepprmint)

table(ppmint_mut$key == ppmint_ref$key)
table(ppmint_mut$sample == ppmint_ref$sample)

g1 = ggplot(ppmint_ref, aes(x=pepprmint)) + xlab("PEPPRMINT score (ref)") + 
  geom_histogram(color="darkblue", fill="lightblue")
g1
```

## Compare PEPPRMINT estimates between reference and mutation
```{r fig.height=3.5, fig.width=4.5}
ppmint = merge(ppmint_mut, ppmint_ref, by = c("key", "sample"), 
               suffixes = c("_mut", "_ref"))
ppmint

gg2 = ggplot(data = ppmint, mapping = aes(x = pepprmint_ref, y = pepprmint_mut)) +
  geom_pointdensity(size=1) + xlab("PEPPRMINT ref") + ylab("PEPPRMINT mut") + 
  scale_color_viridis() + theme(legend.position = "top")

pdf("step4_nb_analysis_files/pepprmint_ref_vs_mut.pdf", width=3.5, height=4.2)
gg2
dev.off()
```

# Read in patient information

```{r}
sample_mb = fread(file = "../data/riaz_patient_mb_info.txt")
dim(sample_mb)
sample_mb[1:2,]

ppmint$Patient = sub("_.*", "", ppmint$sample)
netmhc$Patient = sub("_.*", "", netmhc$sample)
sample_mb$Patient = sub("_.*", "", sample_mb$sample)

clinic = read_excel("../data/_supp/mmc2.xlsx", skip=1)
dim(clinic)
clinic[1:2,]

clinic = as.data.frame(clinic)
table(clinic$Patient %in% sample_mb$Patient)
table(tolower(clinic$Patient) %in% tolower(sample_mb$Patient))

patient = intersect(clinic$Patient, sample_mb$Patient)
length(patient)

patient = intersect(patient, ppmint$Patient)
length(patient)

clinic[which(! clinic$Patient %in% patient),]

clinic = clinic[match(patient, clinic$Patient),]
dim(clinic)
head(clinic)
names(clinic)[c(4:5,7)] = c("Dead", "Time", "Mutation")
head(clinic)
```

# Prepare clinical information

```{r}
table(clinic$Response)
clinic[["Response3"]] = clinic$Response
clinic$Response3[which(clinic$Response == "NE")] = "PD"
clinic$Response3[which(clinic$Response == "CR")] = "PRCR"
clinic$Response3[which(clinic$Response == "PR")] = "PRCR"

clinic$Response = clinic$Response3
clinic$Response[which(clinic$Response3 == "SD")] = "PDSD"
clinic$Response[which(clinic$Response3 == "PD")] = "PDSD"

table(clinic$Response, clinic$Response3)

clinic$Cohort[which(clinic$Cohort == "NIV3-NAIVE")] = "Ipi naive"
clinic$Cohort[which(clinic$Cohort == "NIV3-PROG")]  = "Ipi prog"
```

# Get mutation burden information

```{r}
mat_pre = match(paste0(patient, "_pre"), sample_mb$sample)
clinic[["mb_pre"]] = log10(sample_mb$n_mutations_neoAg[mat_pre] + 1)

mat_on = match(paste0(patient, "_on"), sample_mb$sample)
clinic[["mb_on"]]  = log10(sample_mb$n_mutations_neoAg[mat_on] + 1)
```


# Check the relation between mutation burden and clinical response

```{r}
clinic$mut_diff = clinic$mb_pre - clinic$mb_on
anova(lm(mut_diff ~ Response, data = clinic))
anova(lm(mb_pre ~ Response, data = clinic))

p0 = ggplot(clinic, aes(x=Response, y=mb_pre)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(position=position_jitter(0.2), size=1) + 
  labs(title="", x="Response", y = "log10(MB)") 
p0

p0 = ggplot(clinic, aes(x=Response, y=mb_pre - mb_on)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(position=position_jitter(0.2), size=1) + 
  labs(title="", x="Response", y = "log10(MB) decrease") 
p0

res_cox = coxph(Surv(Time, Dead) ~ mb_pre, data = clinic)
res_cox
```

# Estimate neoAg burden

## Assess association between neoAg versus cytolytic_score

```{r}
cutoffs = seq(0.75, 0.95, by=0.05)
pvals   = matrix(NA, nrow=length(cutoffs), ncol=2)

netmhc$sample = as.factor(netmhc$sample)
ppmint$sample = as.factor(ppmint$sample)

clinic$`Cytolytic Score` = as.numeric(clinic$`Cytolytic Score`)
summary(clinic$cytolytic_score)

g1 = ggplot(clinic, aes(x=log10(`Cytolytic Score`))) + 
  xlab("log10(Cytolytic Score)") + 
  geom_histogram(color="darkblue", fill="lightblue")
g1

clinic[["cytolytic_score"]] = log10(clinic$`Cytolytic Score`)
  
for(i in 1:length(cutoffs)){
  ci = cutoffs[i]
  
  tti       = table(netmhc$sample[netmhc$EL_score_mut_max >= ci])
  mat_pre   = match(paste0(patient, "_pre"), names(tti))
  nb_netmhc_pre = log10(tti[mat_pre] + 1)
  pvals[i,1] = anova(lm(clinic$cytolytic_score ~ nb_netmhc_pre))$`Pr(>F)`[1]

  tti       = table(ppmint$sample[ppmint$pepprmint_mut >= ci])
  mat_pre   = match(paste0(patient, "_pre"), names(tti))
  nb_ppmint_pre = log10(tti[mat_pre] + 1)
  pvals[i,2] = anova(lm(clinic$cytolytic_score ~ nb_ppmint_pre))$`Pr(>F)`[1]
}
pvals

pval_df = data.frame(pval = c(pvals), 
                     cutoff = rep(as.character(cutoffs), times=2),
                     method = rep(c("NetMHCpan-4.1", "PEPPRMINT"), each=5))

pm = ggplot(pval_df, aes(x = method, y = -log10(pval), fill = cutoff)) + 
  geom_bar( stat="identity", position=position_dodge()) + 
  scale_fill_brewer(palette="Paired") + theme(legend.position = "top") +  
  geom_hline(yintercept = -log10(0.05), lty=2) + 
  guides(fill=guide_legend(nrow=2, byrow=TRUE))

pdf("step4_nb_analysis_files/nb_vs_cytolytic_pvals.pdf", width=2.8, height=4)
print(pm)
dev.off()

anova(lm(clinic$cytolytic_score ~ clinic$mb_pre))
```

## Illustrate the relation between neoAg versus cytolytic_score

Here we use 0.9 as cutoff of neoantigen score. 

```{r fig.height=3, fig.width=8}
ci = 0.9

tti       = table(netmhc$sample[netmhc$EL_score_mut_max >= ci])
mat_pre   = match(paste0(patient, "_pre"), names(tti))
nb_netmhc_pre = log10(tti[mat_pre] + 1)

mat_on    = match(paste0(patient, "_on"), names(tti))
nb_netmhc_on  = log10(tti[mat_on] + 1)


tti       = table(ppmint$sample[ppmint$pepprmint_mut >= ci])
mat_pre   = match(paste0(patient, "_pre"), names(tti))
nb_ppmint_pre = log10(tti[mat_pre] + 1)

mat_on    = match(paste0(patient, "_on"), names(tti))
nb_ppmint_on  = log10(tti[mat_on] + 1)

clinic[["nb_netmhc_pre"]] = nb_netmhc_pre
clinic[["nb_ppmint_pre"]] = nb_ppmint_pre

clinic2 = clinic[which(!is.na(clinic$`Cytolytic Score`)),]

lm1  = lm(cytolytic_score ~ mb_pre, data=clinic2)
lm1
pval = anova(lm1)$`Pr(>F)`[1]
g1 = ggplot(clinic2, aes(x=mb_pre, y=cytolytic_score)) +
  geom_point() + xlab("mutation burden\n") + geom_smooth(method=lm) + 
  ggtitle(sprintf("pvalue = %.3f", pval))

pval = anova(lm(cytolytic_score ~ nb_netmhc_pre, data=clinic2))$`Pr(>F)`[1]
g2 = ggplot(clinic2, aes(x=nb_netmhc_pre, y=cytolytic_score)) + 
  geom_point() + xlab("neoantigen burden \n(NetMHCpan 4.1)") + 
  geom_smooth(method=lm) + 
  ggtitle(sprintf("pvalue = %.3f", pval))

pval = anova(lm(cytolytic_score ~ nb_ppmint_pre, data=clinic2))$`Pr(>F)`[1]
g3 = ggplot(clinic2, aes(x=nb_ppmint_pre, y=cytolytic_score)) +
  geom_point() + xlab("neoantigen burden \n(PEPPRMINT)") + 
  geom_smooth(method=lm) + 
  ggtitle(sprintf("pvalue = %.3f", pval))

gg1 = ggarrange(g1, g2, g3, ncol = 3, 
          common.legend = TRUE, legend = "bottom")
gg1
pdf("step4_nb_analysis_files/nb_vs_cytolytic.pdf", width=8, height=3)
print(gg1)
dev.off()
```


## Check the relation between neoantigen change and treatment response

```{r fig.height=3, fig.width=5.6}
ci = 0.9

tti       = table(netmhc$sample[netmhc$EL_score_mut_max >= ci])
mat_pre   = match(paste0(patient, "_pre"), names(tti))
nb_netmhc_pre = log10(as.numeric(tti[mat_pre]) + 1)

mat_on    = match(paste0(patient, "_on"), names(tti))
nb_netmhc_on  = log10(as.numeric(tti[mat_on]) + 1)

tti       = table(ppmint$sample[ppmint$pepprmint_mut >= ci])
mat_pre   = match(paste0(patient, "_pre"), names(tti))
nb_ppmint_pre = log10(as.numeric(tti[mat_pre]) + 1)

mat_on    = match(paste0(patient, "_on"), names(tti))
nb_ppmint_on  = log10(as.numeric(tti[mat_on]) + 1)

clinic[["nb_netmhc_pre"]] = nb_netmhc_pre
clinic[["nb_ppmint_pre"]] = nb_ppmint_pre

clinic[["nb_netmhc_on"]] = nb_netmhc_on
clinic[["nb_ppmint_on"]] = nb_ppmint_on

clinic$mut_diff = clinic$mb_pre - clinic$mb_on
anova(lm(mut_diff ~ Response, data = clinic))

clinic$nb_netmhc_diff = clinic$nb_netmhc_pre - clinic$nb_netmhc_on
anova(lm(nb_netmhc_diff ~ Response, data = clinic))

clinic$nb_ppmint_diff = clinic$nb_ppmint_pre - clinic$nb_ppmint_on
anova(lm(nb_ppmint_diff ~ Response, data = clinic))

table(clinic$Response, is.na(clinic$mut_diff))

table(clinic$Response, is.na(clinic$nb_netmhc_diff))
table(clinic$Response, is.na(clinic$nb_ppmint_diff))
clinic[clinic$Response == "PRCR" & !is.na(clinic$nb_ppmint_diff),]

clinic$Response = factor(clinic$Response, levels = c("PDSD", "PRCR"))

g0 = ggplot(clinic, aes(x = nb_netmhc_pre, y = nb_netmhc_on, color=Response)) +
    geom_jitter(width = 0.02, height=0.02, size = 1) + 
  xlab("pre-therapy") + ylab("on-therapy") + 
  ggtitle("NetMHCpan-4.1 \n neoantigen burden") + xlim(-0.02, 2.2)

g1 = ggplot(clinic, aes(x = nb_ppmint_pre, y = nb_ppmint_on, color=Response)) +
    geom_jitter(width = 0.02, height=0.02, size = 1) + 
  xlab("pre-therapy") + ylab("on-therapy") + 
  ggtitle("PEPPRMINT \n neoantigen burden") + xlim(-0.02, 2.2)

gg2 = ggarrange(g0, g1, ncol = 2, common.legend = TRUE, legend = "bottom", 
                labels = c("(A)", "(B)"))
gg2

pdf("step4_nb_analysis_files/compare_nb_pre_on.pdf", width=5.6, height=3)
print(gg2)
dev.off()
```


# Select potential neoantigens

```{r}
cols = c("sample", "key", "EL_score_mut_max", "EL_score_ref_max", "Patient")
netmhc = netmhc[, ..cols]
median(netmhc$EL_score_ref_max)

table(netmhc$EL_score_mut_max > 0.8 & netmhc$EL_score_ref_max < 0.22)
table(netmhc$EL_score_mut_max > 0.8 & netmhc$EL_score_ref_max < 0.2)
table(netmhc$EL_score_ref_max > 0.8 & netmhc$EL_score_mut_max < 0.2)

table(netmhc$EL_score_mut_max > 0.9 & netmhc$EL_score_ref_max < 0.22)
table(netmhc$EL_score_mut_max > 0.9 & netmhc$EL_score_ref_max < 0.1)
table(netmhc$EL_score_ref_max > 0.9 & netmhc$EL_score_mut_max < 0.1)

cols = c("sample", "key", "pepprmint_mut", "pepprmint_ref", "Patient")
ppmint = ppmint[, ..cols]

median(ppmint$pepprmint_ref)

table(ppmint$pepprmint_mut > 0.8 & ppmint$pepprmint_ref < 0.5)
table(ppmint$pepprmint_mut > 0.8 & ppmint$pepprmint_ref < 0.2)
table(ppmint$pepprmint_ref > 0.8 & ppmint$pepprmint_mut < 0.2)

table(ppmint$pepprmint_mut > 0.9 & ppmint$pepprmint_ref < 0.5)
table(ppmint$pepprmint_mut > 0.9 & ppmint$pepprmint_ref < 0.1)
table(ppmint$pepprmint_ref > 0.9 & ppmint$pepprmint_mut < 0.1)

neoAg_netMHC = netmhc[EL_score_mut_max > 0.9 & EL_score_ref_max < 0.22]
neoAg_ppmint = ppmint[pepprmint_mut > 0.9 & ppmint$pepprmint_ref < 0.5]

dim(neoAg_netMHC)
dim(neoAg_ppmint)

neoAg_netMHC[1:2,]
neoAg_ppmint[1:2,]

t1 = table(neoAg_netMHC$key)
table(t1)

t2 = table(neoAg_ppmint$key)
table(t2)

df_both = merge(neoAg_netMHC, neoAg_ppmint, by=c("sample", "key"))
dim(df_both)
df_both[1:2,]

t1 = table(neoAg_netMHC$sample)
t2 = table(neoAg_ppmint$sample)
table(t1)
table(t2)

sort(t1, decreasing = TRUE)[1:10]
sort(t2, decreasing = TRUE)[1:10]

t1p = t1[t1 > 0]
t2p = t2[t2 > 0]
length(t1p)
length(t2p)
length(intersect(names(t1p), names(t2p)))


t1 = table(neoAg_netMHC$Patient)
t2 = table(neoAg_ppmint$Patient)
table(t1)
table(t2)

sort(t1, decreasing = TRUE)[1:10]
sort(t2, decreasing = TRUE)[1:10]

length(intersect(names(t1), names(t2)))

table(clinic$Patient %in% names(t1), clinic$Response)
table(clinic$Patient %in% names(t2), clinic$Response)

fisher.test(clinic$Patient %in% names(t1), clinic$Response)
fisher.test(clinic$Patient %in% names(t2), clinic$Response)
```

# Add mutation annotation data

```{r}
load("../data/riaz_mutdata.RData")
dim(riaz_mutdata)
names(riaz_mutdata)

riaz_mutdata = riaz_mutdata[,c(1, 9:13, 74:75, 77:78)]
riaz_mutdata$key = paste(gsub("chr", "", riaz_mutdata$seqnames), 
                         riaz_mutdata$start, sep=":")
riaz_mutdata$key = paste(riaz_mutdata$key, 
                         riaz_mutdata$REF, sep=":")
riaz_mutdata$key = paste(riaz_mutdata$key, 
                         riaz_mutdata$ALT, sep=":")
riaz_mutdata[1:2,]

table(neoAg_ppmint$key %in% riaz_mutdata$key)

mat1 = match(neoAg_netMHC$key, riaz_mutdata$key)
neoAg_netMHC_with_anno = cbind(neoAg_netMHC, riaz_mutdata[mat1,-(1:6)])

mat2 = match(neoAg_ppmint$key, riaz_mutdata$key)
neoAg_ppmint_with_anno = cbind(neoAg_ppmint, riaz_mutdata[mat2,-(1:6)])

t1 = table(neoAg_netMHC_with_anno$Gene.ensGene)
table(t1)

sort(t1, decreasing = TRUE)[1:10]


t2 = table(neoAg_ppmint_with_anno$Gene.ensGene)
table(t2)

sort(t2, decreasing = TRUE)[1:10]

fwrite(neoAg_netMHC_with_anno, 
       file="step4_nb_analysis_files/neoAg_netMHC_with_anno.tsv", sep = "\t")

fwrite(neoAg_ppmint_with_anno, 
       file="step4_nb_analysis_files/neoAg_ppmint_with_anno.tsv", sep = "\t")
```

# SessionInfo

```{r}
sessionInfo()
```
