---
title: "TCGA Data Analysis"
output:
  html_document:
    df_print: paged
---

# Load required library
```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(R.utils)
library(ICeDT)
library(ggcorrplot)
```


Read CIBERSORT cell fraction estimates. The output could be txt or csv format. Here we just double check that the results are the same from the two formats. 

```{r}
CBST_folder = "./CIBERSORT/CIBERSORTx_output"

cb_LM22    = read.csv(file.path(CBST_folder,"CIBERSORTx_LM22_Results.csv"))
cb_LM22_v2 = read.table(file.path(CBST_folder,"CIBERSORTx_LM22_Adjusted.txt"), 
                        header=TRUE, as.is=TRUE)
cb_SF2018  = read.csv(file.path(CBST_folder,"CIBERSORTx_SF2018_Results.csv"))
cb_SF2018_filtered = read.csv(file.path(CBST_folder, 
                     "CIBERSORTx_SF2018_filtered_cells_Results.csv"))
dim(cb_LM22)
cb_LM22[1:2,c(1:2,23:26)]

dim(cb_LM22_v2)
cb_LM22_v2[1:2,c(1:2,23:26)]

summary(c(data.matrix(cb_LM22[,-1])) - c(data.matrix(cb_LM22_v2[,-1])))

dim(cb_SF2018)
cb_SF2018_filtered[1:2,c(1:2,12:15)]

dim(cb_SF2018)
cb_SF2018[1:2,c(1:2,12:15)]

summary(cb_LM22$Correlation)
summary(cb_SF2018$Correlation)
summary(cb_SF2018_filtered$Correlation)
```

Boxplot of cell fraction for each cell type inferred using LM22 signature matrix in CIBERSORTx
```{r fig.height=7}
par(mar=c(10,3,1,0.8), mgp=c(1.5,0.5,0), cex.axis=0.8, bty="n")
boxplot(cb_LM22[,2:23],las=2, 
        main="CIBERSORT cell proportion, LM22 signature matrix")
```

Boxplot of cell fraction for each cell type inferred using SF2018 signature matrix in CIBERSORTx
```{r fig.height=5}
par(mar=c(8,3,1,0.8), mgp=c(1.5,0.5,0), cex.axis=0.8, bty="n")
boxplot(cb_SF2018[,2:12], las=2, 
        main="CIBERSORT cell proportion, SF2018 signature matrix")
```

Boxplot of cell fraction for each cell type inferred using SF2018_filtered_cells signature matrix in CIBERSORTx. Here filterd cells means using 200 cells with highest proportion of expressed genes per cell type.
```{r fig.height=5}
par(mar=c(8,3,1,0.8), mgp=c(1.5,0.5,0), cex.axis=0.8, bty="n")
boxplot(cb_SF2018_filtered[,2:12], las=2, 
        main="CIBERSORT cell proportion, SF2018_few0 signature matrix")
```

Correlation plot compare cell fractions inferred using LM22 signature matrix and SF2018 signature matrix in CIBERSORTx            
- For most of the cell types, we see the fractions inferred using different signature matrix correlated
- Fractions of dendritic cell inferred using the 2 types of signature matrices have small correlation
- The fractions three types of NK cells also showed some dubious correlation: such as correlation between NK1 and Macrophages.M0. 
```{r fig.height=5, fig.width=8, warning=FALSE, message = FALSE}

table(cb_LM22$Mixture == cb_SF2018$Mixture)
cb_LM22_mtrx   = data.matrix(cb_LM22[,2:23])
cb_SF2018_mtrx = data.matrix(cb_SF2018[,2:12])
cb_SF2018_filtered_mtrx = data.matrix(cb_SF2018_filtered[,2:12])

cormat_LM22_SF2018 = cor(cb_LM22_mtrx, cb_SF2018_mtrx, 
                         method = "spearman")
cormat_LM22_SF2018_filtered = cor(cb_LM22_mtrx, 
                                  cb_SF2018_filtered_mtrx, 
                         method = "spearman")

summary(c(cormat_LM22_SF2018))
summary(c(cormat_LM22_SF2018_filtered))

ggcorrplot(t(cormat_LM22_SF2018), tl.cex = 6) + 
  scale_fill_gradient2(limit = c(-0.8, 0.8), low = "blue", 
                       high =  "red", mid = "white") 

round(cormat_LM22_SF2018,2)

ggcorrplot(t(cormat_LM22_SF2018_filtered), tl.cex = 6) + 
    scale_fill_gradient2(limit = c(-0.8, 0.8), low = "blue", 
                       high =  "red", mid = "white") 

round(cormat_LM22_SF2018_filtered,2)

```

Correlation plot compare cell fractions inferred using SF2018 signature matrix and SF2018_filtered signature matrix in CIBERSORTx         
- We saw positive correlation for the fraction of the same cell type inferred by the 2 different signature matrix. 
```{r warning=FALSE, message = FALSE}

cormat_SF2018 = cor(cb_SF2018_mtrx, cb_SF2018_filtered_mtrx, 
                         method = "spearman")

summary(c(cormat_SF2018))

ggcorrplot(t(cormat_SF2018), tl.cex = 6) 
```

```{r}
par(mar=c(3,3,1,0.8), mgp=c(1.5,0.5,0), cex.axis=0.85, cex.lab=0.85, cex.main=0.85)
layout(matrix(c(1,2,3,4),nrow=2,ncol = 2,byrow = TRUE))

plot(cb_SF2018$CD8T_B, cb_LM22$T.cells.CD8, 
     xlab = "SF2018 CD8 B cell", ylab = "LM22 CD8 T cell")

plot(cb_SF2018$CD8T_G, cb_LM22$T.cells.CD8, 
     xlab = "SF2018 CD8 G cell", ylab = "LM22 CD8 T cell")

plot(cb_SF2018_filtered$CD8T_B, cb_LM22$T.cells.CD8, 
     xlab = "SF2018_few0 CD8 B cell", ylab = "LM22 CD8 T cell")

plot(cb_SF2018_filtered$CD8T_G, cb_LM22$T.cells.CD8, 
     xlab = "SF2018_few0 CD8 G cell", ylab = "LM22 CD8 T cell")
```

```{r}
plot(cb_SF2018$CD8T_B, cb_SF2018$CD8T_G, 
     xlab = "SF2018 CD8 B cell", ylab = "SF2018 CD8 G cell")
```

```{r}
plot(cb_SF2018_filtered$CD8T_B, cb_SF2018_filtered$CD8T_G, 
     xlab = "cb_SF2018_filtered CD8 B cell", 
     ylab = "cb_SF2018_filtered CD8 G cell")
```


```{r}
gc()
sessionInfo()
```
