---
title: "TCGA Data Analysis"
output:
  html_document:
    df_print: paged
---

# Installing the ICeDT library
```{r eval=FALSE}
library(devtools)
install_github("Sun-lab/ICeDT")
```

# Load required library
```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(R.utils)
library(ICeDT)
```

# Data path and information datasets
```{r}
TCGAfolder = "../../TCGA/"
datafolder = paste0(TCGAfolder,"SKCM/gdc_download_20201012_225902.677830/")

# Info matrix on the samples
sampleinfo = read.table(paste0(TCGAfolder,"SKCM/gdc_sample_sheet.2020-10-12.tsv"), header = TRUE, sep = "\t", as.is = TRUE)
head(sampleinfo)
# data category: Transcriptome Profiling
# data type: Gene Expression Quantification
# Project ID: TCGA-SKCM
# Case ID: 468 case ID, not unique for all samples
# Sample ID: 472 sample ID, unique for all samples
# Sample type: "Metastatic", "Primary Tumor", "Additional Metastatic", "Solid Tissue Normal" 

# Info matrix on the genes: 60483 genes
geneinfo = fread(paste0(TCGAfolder, "gencode.v22.genes.txt"), header = TRUE, sep = "\t", drop = 8, na.strings = c("NA", ""))
head(geneinfo)

# Info matrix on the length of each gene: 60483 genes
load(paste0(TCGAfolder,"Gene_Lengths.RData"))
head(GeneLengths.Mat) # 60483 * 3

# LM22 signature matrix
LM22_sigmat = read.table("../../_reference/LM22.txt",header = TRUE, sep = "\t", as.is = TRUE)
LM22_sigmat[1:5,1:5] # 547*23
```

# Step 1: Prepare count matrix
- read in each count data
- match the filename of count data to the sample ID
- check the gene ID of each new file is the same with the previous files
- combine the count data of all samples to a single count matrix
```{r}
# file paths for all individual count files
countfiles = paste0(datafolder,sampleinfo$File.ID, "/",sampleinfo$File.Name)
# save the gene names in the count files
gene.id = NA
# save the gene counts for each sample, col.names are sample ID
count.df = data.frame(matrix(numeric(), 0, length(countfiles),
                dimnames=list(c(), sampleinfo$Sample.ID)))

for (f in countfiles) {
  # read in the count file
  cf = fread(f, col.names = c("ensembl.ID", "counts"))
  # when reading in the first count file, save the gene names to gene.id, 
  # and save the counts of this sample to the corresponding column in the count matrix
  if (anyNA(gene.id)){
    gene.id = cf$ensembl.ID
    count.df[1:length(cf$counts),colnames(count.df)[match(f,countfiles)]] = cf$counts
  } else {
    # starting from the second file, first check if its gene names match those in the first file
    # if matching, save the counts of this sample to the corresponding column in the count matrix
    if (all(gene.id == cf$ensembl.ID)){
      count.df[1:length(cf$counts),colnames(count.df)[match(f,countfiles)]] = cf$counts
    } else {
      # if not matching, print error message, skip this sample
      cat("wrong ensembl id in ", f, "\n")
      next
    }
  }
}
rownames(count.df) = gene.id
count.df[1:5,1:5]
rm(cf)

# sanity check
anyNA(count.df)
dim(count.df) # 60488 * 472

# Count.df has 60488 rows:
# - the last 5 rows are QC statistics, the others are counts for genes. 
# Match the rows of count.df to the rows of geneinfo by ensembl ID: 
# - 60483 gene ensembl ID in the data 
# - all matched to those in the gene info matrix in the same order
count.df = count.df[match(geneinfo$geneId, gene.id),] # 60483 * 472
gene.id = row.names(count.df)

# save the raw matrix of count
write.csv(count.df, "raw_counts.csv")
```

# Step 2: Match gene ensembl ID to gene names
```{r}
# Some ensembl ID are missing hgnc symbols
missing.id = which(is.na(geneinfo$hgnc_symbol))
geneinfo.nomissing = geneinfo[-missing.id,]
# Some ensembl ID have duplicated hgnc symbols
duplicated.id = which(duplicated(geneinfo.nomissing$hgnc_symbol))
geneinfo.filtered = geneinfo.nomissing[-duplicated.id,]

# Update count.df to only genes with 1-1 mapping between their ensembl ID and hgnc symbols, 
# 35295 genes remained in the end. 
count.df.filtered = cbind(ensembl_ID = as.character(geneinfo.filtered$geneId), 
                          hgnc_symbol = as.character(geneinfo.filtered$hgnc_symbol), 
                          count.df[match(geneinfo.filtered$geneId, gene.id),])
rownames(count.df.filtered) = NULL
count.df.filtered[1:5,1:5]

# Update count.df to only genes intersect between TCGA data and LM22 data
## 535 genes in common between the LM22 matrix and our count matrix
reference.genes.id = na.omit(match(LM22_sigmat$Gene.symbol, count.df.filtered$hgnc_symbol)) 
count.df.refgenes = count.df.filtered[reference.genes.id,]
count.df.refgenes[1:5,1:5]

rm(geneinfo.nomissing)
rm(geneinfo.filtered)
```

Save the count matrix     
```{r eval=FALSE}
write.table(count.df.filtered, "count_matrix.txt", quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
write.table(count.df.refgenes, "count_matrix_reference_genes.txt", quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
```

# Step 3: Transform count data to TPM data
```{r}
calculate_TPM = function(count, gene_length, indices) {
  if (nrow(count) != length(gene_length)) stop("Number of rows of the count matrix does not match gene lengths!")
  TPM = count / gene_length
  t(t(TPM)*1e6/colSums(TPM[indices, ]))
}

geneLength = GeneLengths.Mat$Exonic[match(count.df.refgenes$ensembl_ID, GeneLengths.Mat$Gencode.ID)]
SKCM_TPM = calculate_TPM(count.df.refgenes[,c(-1,-2)], geneLength, 1:nrow(count.df.refgenes)) 

rownames(SKCM_TPM) = count.df.refgenes$hgnc_symbol
dim(SKCM_TPM) # 535 * 472
SKCM_TPM[1:5,1:3]
```

# Generate CIBERSORT input

```{r}
# observed TPM data
geneSymbol_and_observed_TPM = cbind(geneSymbol = count.df.refgenes$hgnc_symbol, SKCM_TPM)
# reference gene expression data
signature_matrix = as.matrix(LM22_sigmat[match(rownames(SKCM_TPM), LM22_sigmat$Gene.symbol),-1])
rownames(signature_matrix) = rownames(SKCM_TPM)
geneSymbol_and_signature_matrix = cbind(geneSymbol = count.df.refgenes$hgnc_symbol, signature_matrix)
```

```{r eval=FALSE}
CBSTfolder = "./CIBERSORT"
signature_gene_file = file.path(CBSTfolder,"CIBERSORT_input_signature_gene_SKCM.txt")
mixture_file = file.path(CBSTfolder,"CIBERSORT_input_observed_TPM_SKCM.txt")

# generate matrices as input for CIBERSORT online
write.table(geneSymbol_and_observed_TPM,
            file = mixture_file,
            quote = FALSE,
            row.names = FALSE,
            sep="\t")
write.table(geneSymbol_and_signature_matrix,
            file = signature_gene_file,
            quote = FALSE,
            row.names = FALSE,
            sep="\t")
```

CIBERSORT output
```{r}
CBSTfolder = "./CIBERSORT"
cibersort_cell_fraction = read.table(file.path(CBSTfolder,"CIBERSORTx_output/CIBERSORTxGEP_Job2_Fractions.txt"), header = TRUE, sep = "\t")
head(cibersort_cell_fraction)
```

# Run ICeDT
Got error at this step: likelihood decreased during EM
```{r eval=FALSE}
icedt_output_file =  "SKCM_ICeDT_output.rds"
if (!file.exists(icedt_output_file)) {
  set.seed(34548)
  icedt_output = ICeDT::ICeDT(
    Y = SKCM_TPM,
    Z = signature_matrix,
    tumorPurity = NULL,
    refVar = NULL)
  # save to ICeDT cellular frequency RDS file
  saveRDS(icedt_output, icedt_output_file)
} else {
  icedt_output = readRDS(icedt_output_file)
}
```


```{r}
gc()
sessionInfo()
```
