---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(Rtsne)
library(R.utils)
library(ggplot2)
```

# Processing data

```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
rm(sf_tpm1)
rm(sf_tpm2)
```

```{r}
dim(sf_tpm) # 55737 x 16292
cell.names = colnames(sf_tpm)[-1]
gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

# Initial QC

## How many transcripts per cell
```{r}
grouping = rep(1:82, each=200)[1:length(cell.names)]
reads.per.cell = NULL
for (i in 1:82) {
  ss = cell.names[grouping == i]
  temp.reads = unlist(sf_tpm[, lapply(.SD, sum), .SDcols = ss])
  reads.per.cell = c(reads.per.cell,temp.reads)
}
hist(reads.per.cell, main = "Distribution of the amount of transcripts per cell", xlab = "reads per cell")
```

## How many genes are expressed per cell
```{r}
genes.per.cell = NULL
for (i in 1:82) {
  ss = cell.names[grouping == i]
  temp.counts = unlist(sf_tpm[, lapply(.SD, FUN = function(x) sum(x>0)), .SDcols = ss])
  genes.per.cell = c(genes.per.cell,temp.counts)
}
hist(genes.per.cell, main = "Distribution of the number of expressed genes per cell", xlab = "number of espressed genes per cell")
rm(grouping)
```

## How many cells each gene is expressed  
```{r}
gene.grouping = rep(1:28, each = 2000)[1:dim(sf_tpm)[1]]
cells.per.gene = NULL
for (i in 1:28) {
  temp.counts = apply(sf_tpm[gene.grouping==i,], 1, FUN = function(x) sum(x>0))
  cells.per.gene = c(cells.per.gene,temp.counts)
}
hist(log10(cells.per.gene), main = "Distribution of number of cells each genes expressed", xlab = "log10(number of cells) each gene expressed")
rm(gene.grouping)
```

# Choose high-variance genes

```{r}
# kept 12705 genes
large.express.genes = gene.names[which(cells.per.gene > 500)]
large.expressed.tpm = sf_tpm[match(large.express.genes,rownames(sf_tpm)),]
# plot the variance of the expression levels for each gene against its mean expression level
genes.mean = apply(large.expressed.tpm, 1, mean)
genes.variance = apply(large.expressed.tpm, 1, var)
df_mv = data.frame(gene=large.express.genes, mean=genes.mean, var=genes.variance)
rm(large.express.genes)
rm(large.expressed.tpm)
rm(genes.mean)
rm(genes.variance)
ggplot(data = df_mv) + 
  geom_point(aes(x=mean,y=var),size = 0.1)
```

```{r}
# divide the genes into 20 equal-size bins based on their mean expression levels
df_mv$grouping = cut_number(df_mv$mean, n=20)
table(df_mv$grouping)
# in each bin, select the first 50 genes with the highest variance
hv.genes.list = by(df_mv, df_mv$grouping, FUN = function(df) return(df$gene[order(df$var,decreasing = TRUE)][1:50]))
hv.genes = as.character(unlist(hv.genes.list))
rm(hv.genes.list)
hvg_tpm = sf_tpm[match(hv.genes,rownames(sf_tpm)),]
hvg_tpm = t(hvg_tpm)
colnames(hvg_tpm) = hv.genes
rm(df_mv)
```

## Perform PCA on selected genes

```{r}
pca.hvg = princomp(hvg_tpm, cor = TRUE)
rm(hvg_tpm)
```

### screen plot
```{r}
eigvals = pca.hvg$sdev^2
pve = eigvals/sum(eigvals)
plot(1:50,pve[1:50],main = "scree plot (first 50 PCs)",type="b",xlab="i-th PC",ylab="Proportion of Variance Explained")
rm(eigvals)
```

### plot of cumulative proportion of variance explained
```{r}
yvec = cumsum(pve)
plot(yvec, type="l", main = "plot of cumulative proportion of variance", xlab = "i-th Principal Components", ylab = "Cumulative variance",ylim=c(0,1))
abline(h=0.8, lty = 2)
rm(yvec)
rm(pve)
```

```{r}
pca.scores = pca.hvg$scores
rm(pca.hvg)
```

## Clustering on top 50 PCs

```{r}
top50pcs = pca.scores[,1:50]
rm(pca.scores)
#head(top50pcs)
```

### Calculate TSNE

```{r}
set.seed(9999)
tsne = Rtsne(top50pcs, pca=FALSE)
df_tsne = as.data.frame(tsne$Y)
names(df_tsne) = paste0("topPC_TSNE",seq(ncol(tsne$Y)))
rm(tsne)
```

### K-means clustering on top 50 PCs

Prepare dataframe for k-means clustering using top 50 PCs
```{r}
df_top50PCs = cbind(top50pcs, df_tsne)
rm(df_tsne)
#head(df_top50PCs)
```

Try Kmeans for 6 to 12 clusters 
```{r}
set.seed(9999)
all_num_clust = c(6:12)
for (num_clust in all_num_clust) {
  cat(paste0("KM with ",num_clust," clusters.\n"))
  kmeans_out = kmeans(top50pcs, centers = num_clust, iter.max = 1e3, nstart = 50, algorithm = "MacQueen")
  print(kmeans_out[c("betweenss","tot.withinss")])
  km_label = paste0("KM_",num_clust)
  df_top50PCs[[km_label]] = as.factor(kmeans_out$cluster)
  #pt = ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=eval(as.name(paste(km_label)))),size = 0.1)+scale_colour_discrete(name=paste(km_label))+guides(color = guide_legend(override.aes = list(size = 2)))
  #print(pt)
  rm(kmeans_out)
  #rm(pt)
}
```

Compared to the clustering result from Sade-Feldman et al., 2018

```{r}
xnm = "1-s2.0-S0092867418313941-mmc1.xlsx"
table_s1 = readxl::read_xlsx(paste0("../_reference/Sade_Feldman_2018/", xnm), sheet = "Cluster annotation-Fig1B-C", col_names = TRUE)
df_top50PCs$sf_cluster = as.factor(table_s1$`Cluster number`)
```

Cluster annotation from Sade-Feldman et al., 2018, 11 clusters
```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=sf_cluster),size = 0.1)+guides(color = guide_legend(override.aes = list(size = 2)))
```

Cluster annotation from my Kmeans analysis, 11 clusters
```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=KM_11),size = 0.1)+guides(color = guide_legend(override.aes = list(size = 2)))
```

For the K-mean analysis results using 11 clusters, it seems that using top 20 PCs gives a better result than using top 50 PCs, because the result using 50 PCs didn't identify the cluster of G10 Memory T-cells. Moreover, the G3 Monocytes/Macrophages group was split into 3 clusters. 

```{r}
table(df_top50PCs$KM_11, df_top50PCs$sf_cluster)
```

Cluster annotation from my Kmeans analysis, 12 clusters
```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=KM_12),size = 0.1)+guides(color = guide_legend(override.aes = list(size = 2)))
```

Here the cluster of G10 Memory T-cells is identified, but the result doesn't seem to be too different than the result obtained based on only the top 20 PCs. 

```{r}
table(df_top50PCs$KM_12, df_top50PCs$sf_cluster)
```

### Re-cluster T and NK cells

First, I get all the T and NK cells, corresponding to the cells in clusters 5-11 in Sade_Feldman et al. (2018). There are 12850 such cells. Then, I save the PC scores, t-SNE, and clustering labels defined by Sade_Feldman et al. (2018) of these cells into the dataframe `df_top50PCs_TNKcells`. 

```{r}
T.NK.cells = rownames(df_top50PCs)[as.numeric(df_top50PCs$sf_cluster)>=5]
df_top50PCs_TNKcells = df_top50PCs[match(T.NK.cells, rownames(df_top50PCs)), c(1:52,60)]
```

Next, I tried Kmeans analysis using 6 to 8 clusters, and compared my results to the cluster annotation from the paper
```{r}
set.seed(9999)
all_num_clust = c(6:8)
for (num_clust in all_num_clust) {
  cat(paste0("KM with ",num_clust," clusters.\n"))
  kmeans_out = kmeans(df_top50PCs_TNKcells[,1:50], centers = num_clust, iter.max = 1e3, nstart = 50, algorithm = "MacQueen")
  print(kmeans_out[c("betweenss","tot.withinss")])
  km_label = paste0("KM_",num_clust)
  df_top50PCs_TNKcells[[km_label]] = as.factor(kmeans_out$cluster)
  pt = ggplot(data=df_top50PCs_TNKcells)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=eval(as.name(paste(km_label)))),size = 0.1)+scale_colour_discrete(name=paste(km_label))+guides(color = guide_legend(override.aes = list(size = 2)))
  print(pt)
  rm(kmeans_out)
  rm(pt)
}
```

T and NK cells cluster annotation from Sade-Feldman et al., 2018, clusters 5 - 11.
```{r}
ggplot(data=df_top50PCs_TNKcells)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=sf_cluster),size = 0.1)+guides(color = guide_legend(override.aes = list(size = 2)))
```

Using 8 clusters in K-means clustering gives the best result. Clusters 7, 10, 11 can be identified as single groups. Compare to the clustering using all cells, the improvement is that most cells in clusters 5 and 8 could also be distinguished now. The remaining issue is that clusters 6 and 9 still cannot be separated, and some cells in clusters 5 and 8 still overlapped.

```{r}
table(df_top50PCs_TNKcells$KM_8, df_top50PCs_TNKcells$sf_cluster)
```

```{r}
rm(df_top50PCs_TNKcells)
rm(df_top50PCs)
rm(top50pcs)
rm(T.NK.cells)
```

```{r}
gc()
sessionInfo()
```

