---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(Rtsne)
library(R.utils)
library(ggplot2)
library(ggcorrplot)
```

# Processing data

```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
rm(sf_tpm1)
rm(sf_tpm2)
```

```{r}
dim(sf_tpm) # 55737 x 16292
cell.names = colnames(sf_tpm)[-1]
gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

# Initial QC

## How many transcripts per cell
```{r}
grouping = rep(1:82, each=200)[1:length(cell.names)]
reads.per.cell = NULL
for (i in 1:82) {
  ss = cell.names[grouping == i]
  temp.reads = unlist(sf_tpm[, lapply(.SD, sum), .SDcols = ss])
  reads.per.cell = c(reads.per.cell,temp.reads)
}
hist(reads.per.cell, main = "Distribution of the amount of transcripts per cell", xlab = "reads per cell")
```

## How many genes are expressed per cell
```{r}
genes.per.cell = NULL
for (i in 1:82) {
  ss = cell.names[grouping == i]
  temp.counts = unlist(sf_tpm[, lapply(.SD, FUN = function(x) sum(x>0)), .SDcols = ss])
  genes.per.cell = c(genes.per.cell,temp.counts)
}
hist(genes.per.cell, main = "Distribution of the number of expressed genes per cell", xlab = "number of espressed genes per cell")
rm(grouping)
```

## How many cells each gene is expressed  
```{r}
gene.grouping = rep(1:28, each = 2000)[1:dim(sf_tpm)[1]]
cells.per.gene = NULL
for (i in 1:28) {
  temp.counts = apply(sf_tpm[gene.grouping==i,], 1, FUN = function(x) sum(x>0))
  cells.per.gene = c(cells.per.gene,temp.counts)
}
hist(log10(cells.per.gene), main = "Distribution of number of cells each genes expressed", xlab = "log10(number of cells) each gene expressed")
rm(gene.grouping)
```

# Choose high-variance genes

```{r}
# kept 12705 genes
large.express.genes.all = gene.names[which(cells.per.gene > 500)]
large.expressed.tpm = sf_tpm[match(large.express.genes.all,rownames(sf_tpm)),]
# plot the variance of the expression levels for each gene against its mean expression level
genes.mean = apply(large.expressed.tpm, 1, mean)
genes.variance = apply(large.expressed.tpm, 1, var)
df_mv = data.frame(gene=large.express.genes.all, mean=genes.mean, var=genes.variance)
rm(large.expressed.tpm)
rm(genes.mean)
rm(genes.variance)
ggplot(data = df_mv) + 
  geom_point(aes(x=mean,y=var),size = 0.1)
```

```{r}
# divide the genes into 20 equal-size bins based on their mean expression levels
df_mv$grouping = cut_number(df_mv$mean, n=20)
table(df_mv$grouping)
# in each bin, select the first 50 genes with the highest variance
hv.genes.list = by(df_mv, df_mv$grouping, FUN = function(df) return(df$gene[order(df$var,decreasing = TRUE)][1:50]))
hv.genes.all = as.character(unlist(hv.genes.list))
rm(hv.genes.list)
hvg_tpm = sf_tpm[match(hv.genes.all,rownames(sf_tpm)),]
hvg_tpm = t(hvg_tpm)
colnames(hvg_tpm) = hv.genes.all
rm(df_mv)
```

## Perform PCA on selected genes

```{r}
pca.hvg = princomp(hvg_tpm, cor = TRUE)
rm(hvg_tpm)
```

### screen plot
```{r}
eigvals = pca.hvg$sdev^2
pve = eigvals/sum(eigvals)
plot(1:50,pve[1:50],main = "scree plot (first 50 PCs)",type="b",xlab="i-th PC",ylab="Proportion of Variance Explained")
rm(eigvals)
```

### plot of cumulative proportion of variance explained
```{r}
yvec = cumsum(pve)
plot(yvec, type="l", main = "plot of cumulative proportion of variance", xlab = "i-th Principal Components", ylab = "Cumulative variance",ylim=c(0,1))
abline(h=0.8, lty = 2)
rm(yvec)
rm(pve)
```

```{r}
pca.scores = pca.hvg$scores
rm(pca.hvg)
```

# Clustering on top 50 PCs

```{r}
top50pcs = pca.scores[,1:50]
rm(pca.scores)
#head(top50pcs)
```

## Calculate TSNE

```{r}
set.seed(9999)
tsne = Rtsne(top50pcs, pca=FALSE)
df_tsne = as.data.frame(tsne$Y)
names(df_tsne) = paste0("topPC_TSNE",seq(ncol(tsne$Y)))
rm(tsne)
```

## K-means clustering on top 50 PCs

Prepare dataframe for k-means clustering using top 50 PCs
```{r}
df_top50PCs = cbind(top50pcs, df_tsne)
rm(df_tsne)
#head(df_top50PCs)
```

Try Kmeans for 6 to 12 clusters 
```{r}
set.seed(9999)
all_num_clust = c(6:12)
for (num_clust in all_num_clust) {
  cat(paste0("KM with ",num_clust," clusters.\n"))
  kmeans_out = kmeans(top50pcs, centers = num_clust, iter.max = 1e3, nstart = 50, algorithm = "MacQueen")
  print(kmeans_out[c("betweenss","tot.withinss")])
  km_label = paste0("KM_",num_clust)
  df_top50PCs[[km_label]] = as.factor(kmeans_out$cluster)
  pt = ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=eval(as.name(paste(km_label)))),size = 0.1)+scale_colour_discrete(name=paste(km_label))+guides(color = guide_legend(override.aes = list(size = 2)))
  print(pt)
  rm(kmeans_out)
  rm(pt)
}
```

Compared to the clustering result from Sade-Feldman et al., 2018

```{r}
xnm = "1-s2.0-S0092867418313941-mmc1.xlsx"
table_s1 = readxl::read_xlsx(paste0("../_reference/Sade_Feldman_2018/", xnm), sheet = "Cluster annotation-Fig1B-C", col_names = TRUE)
df_top50PCs$sf_cluster = as.factor(table_s1$`Cluster number`)
```

Cluster annotation from Sade-Feldman et al., 2018, 11 clusters
```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=sf_cluster),size = 0.1)+guides(color = guide_legend(override.aes = list(size = 2)))
```

Cluster annotation from my Kmeans analysis, 12 clusters
```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=KM_12),size = 0.1)+guides(color = guide_legend(override.aes = list(size = 2)))
```

```{r}
table(df_top50PCs$KM_12, df_top50PCs$sf_cluster)
```

## Select G1 B cells, G2 Plasma cells, G3 Monocytes/Macrophages, G4 Dendric cells, G7 Regulatory T cells and G10 Memory T-cells by taking the largest intersection between the clusters defined in the paper and our clusters 
 
- 1421 B cells: 1
- 301 Plasma cells: 2
- 1342 Monocytes/Macrophages: 3
- 276 Dendric cells: 4
- 1332 CD4 T regulatory cells: 5
- 1451 CD4 T memory cells: 6

```{r}
cell_type = data.frame(cell = cell.names, type = NA)
cell_type$type[df_top50PCs$KM_12==12 & df_top50PCs$sf_cluster == 1] = 1
cell_type$type[df_top50PCs$KM_12==3 & df_top50PCs$sf_cluster == 2] = 2
cell_type$type[(df_top50PCs$KM_12==5 | df_top50PCs$KM_12==7 | df_top50PCs$KM_12==10) & df_top50PCs$sf_cluster == 3] = 3
cell_type$type[df_top50PCs$KM_12==11 & df_top50PCs$sf_cluster == 4] = 4
cell_type$type[df_top50PCs$KM_12==4 & df_top50PCs$sf_cluster == 7] = 5
cell_type$type[df_top50PCs$KM_12==8 & df_top50PCs$sf_cluster == 10] = 6

table(cell_type$type, df_top50PCs$sf_cluster)
```

```{r}
rm(top50pcs)
rm(df_top50PCs)
```

## Select CD8_B cells and CD8_G cells 
 
- 2482 CD8_B cells: 7
- 3043 CD8_G cells: 8

```{r}
# Cell cluster annotation of CD8+ T cells 
table_s2 = read.csv("CD8_cluster.csv", stringsAsFactors = FALSE)
table_s2$ruoyi_cluster[is.na(table_s2$ruoyi_cluster)] = "NA"
CD8T.cells = table_s2$Cell.Name

CD8_B = table_s2$Cell.Name[table_s2$ruoyi_cluster=="CD8_B"]
CD8_G = table_s2$Cell.Name[table_s2$ruoyi_cluster=="CD8_G"]
cell_type$type[match(CD8_B,cell_type$cell)] = 7
cell_type$type[match(CD8_G,cell_type$cell)] = 8
#NK = rownames(df_top50PCs_CDNKcells)[df_top50PCs_CDNKcells$ruoyi_CD8 == "NA"]
table(cell_type$type)
```

# Fine-clustering for NK cells

First, I get all the CD8 and NK cells, corresponding to the cells in clusters 5, 6, 8, 9 and 11 in Sade_Feldman et al. (2018). There are 9337 such cells. Then, I excluded the CD8 T cells using the annotation in the paper. 3513 cells remained after excluding the CD8 T cells. 

## Data for only CD8 and NK cells
```{r}
CD8.NK.cells = cell.names[as.numeric(table_s1$`Cluster number`) %in% c(5,6,8,9,11)]

TNK.cells = CD8.NK.cells[! CD8.NK.cells %in% CD8T.cells] # cells that are not CD8 cells in cluster 5,6,8,9,11
TNKcells_tpm = sf_tpm[, ..TNK.cells]
row.names(TNKcells_tpm) = gene.names
```

## How many cells each gene is expressed  
```{r}
gene.grouping = rep(1:28, each = 2000)[1:dim(TNKcells_tpm)[1]]
cells.per.gene = NULL
for (i in 1:28) {
  temp.counts = apply(TNKcells_tpm[gene.grouping==i,], 1, FUN = function(x) sum(x>0))
  cells.per.gene = c(cells.per.gene,temp.counts)
}
hist(log10(cells.per.gene), main = "Distribution of number of cells each genes expressed", xlab = "log10(number of cells) each gene expressed")
rm(gene.grouping)
```

## Choose high-variance genes

```{r}
# kept 10925 genes
large.express.genes.TNK = gene.names[which(cells.per.gene > 150)]
large.expressed.tpm = TNKcells_tpm[match(large.express.genes.TNK,rownames(TNKcells_tpm)),]
# plot the variance of the expression levels for each gene against its mean expression level
genes.mean = apply(large.expressed.tpm, 1, mean)
genes.variance = apply(large.expressed.tpm, 1, var)
df_mv = data.frame(gene=large.express.genes.TNK, mean=genes.mean, var=genes.variance)
#rm(large.express.genes.TNK)
rm(large.expressed.tpm)
rm(genes.mean)
rm(genes.variance)
ggplot(data = df_mv) + 
  geom_point(aes(x=mean,y=var),size = 0.1)
```

```{r}
# divide the genes into 20 equal-size bins based on their mean expression levels
df_mv$grouping = cut_number(df_mv$mean, n=20)
table(df_mv$grouping)
# in each bin, select the first 50 genes with the highest variance
hv.genes.list = by(df_mv, df_mv$grouping, FUN = function(df) return(df$gene[order(df$var,decreasing = TRUE)][1:50]))
hv.genes.TNK = as.character(unlist(hv.genes.list))
rm(hv.genes.list)
hvg_TNKcells_tpm = TNKcells_tpm[match(hv.genes.TNK,rownames(TNKcells_tpm)),]
hvg_TNKcells_tpm = t(hvg_TNKcells_tpm)
colnames(hvg_TNKcells_tpm) = hv.genes.TNK
rm(df_mv)
```

## Perform PCA on selected genes

```{r}
pca.hvg = princomp(hvg_TNKcells_tpm, cor = TRUE)
rm(hvg_TNKcells_tpm)
pca.scores = pca.hvg$scores
rm(pca.hvg)
```

## Clustering on top 50 PCs

```{r}
top50pcs_TNKcells = pca.scores[,1:50]
rm(pca.scores)
#head(top50pcs_CDNKcells)
```

### Calculate TSNE

```{r}
set.seed(9999)
tsne = Rtsne(top50pcs_TNKcells, pca=FALSE)
df_tsne = as.data.frame(tsne$Y)
names(df_tsne) = paste0("topPC_TSNE",seq(ncol(tsne$Y)))
rm(tsne)
```

### K-means clustering on top 50 PCs

Prepare dataframe for k-means clustering using top 50 PCs
```{r}
df_top50PCs_TNKcells = cbind(top50pcs_TNKcells, df_tsne)
rm(df_tsne)
#head(df_top50PCs_CDNKcells)
```

```{r}
set.seed(9999)
all_num_clust = c(2:6)
for (num_clust in all_num_clust) {
  cat(paste0("KM with ",num_clust," clusters.\n"))
  kmeans_out = kmeans(top50pcs_TNKcells, centers = num_clust, iter.max = 1e3, nstart = 50, algorithm = "MacQueen")
  print(kmeans_out[c("betweenss","tot.withinss")])
  km_label = paste0("KM_",num_clust)
  df_top50PCs_TNKcells[[km_label]] = as.factor(kmeans_out$cluster)
  pt = ggplot(data=df_top50PCs_TNKcells)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=eval(as.name(paste(km_label)))),size = 0.1)+scale_colour_discrete(name=paste(km_label))+guides(color = guide_legend(override.aes = list(size = 2)))
  print(pt)
  rm(kmeans_out)
  rm(pt)
}
```

### Cell cluster annotation from Sade-Feldman et al., 2018   

```{r}
df_top50PCs_TNKcells$sf_cluster = as.factor(table_s1$`Cluster number`[match(TNK.cells, cell.names)])

ggplot(data=df_top50PCs_TNKcells)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=sf_cluster),size = 0.1)+guides(color = guide_legend(override.aes = list(size = 2)))
```

The clustering results from k-means analysis using 3-6 clusters are generally very similar:
G5 and G8 form one cluster, G6 and G9 form one cluster, and G11 forms an individual cluster.   
Using 3 clusters would suffice.

## Obtain the labeling of NK cells

```{r}
NK1 = row.names(df_top50PCs_TNKcells)[df_top50PCs_TNKcells$KM_3==1]
NK2 = row.names(df_top50PCs_TNKcells)[df_top50PCs_TNKcells$KM_3==2]
NK3 = row.names(df_top50PCs_TNKcells)[df_top50PCs_TNKcells$KM_3==3]

cell_type$type[match(NK1,cell_type$cell)] = 9
cell_type$type[match(NK2,cell_type$cell)] = 10
cell_type$type[match(NK3,cell_type$cell)] = 11

table(cell_type$type)
```


## Save the current labeling of the cells that we obtained from the K-means analysis

- 1421 B cells: 1
- 290 Plasma cells: 2
- 1341 Monocytes/Macrophages: 3
- 275 Dendric cells: 4
- 1313 CD4 T regulatory cells: 5
- 1099 CD4 T memory cells: 6
- 2482 CD8_B cells: 7
- 3043 CD8_G cells: 8
- 463 NK1 cells: 9
- 1903 NK2 cells: 10
- 1147 NK3 cells: 11
- 1514 Remaining unclustered cells: 0

```{r eval=FALSE}
write.table(cell_type, "cell_type.txt", quote = FALSE, col.names = TRUE, row.names = FALSE, na = "0")
```

## Compare expression data with CIBERSORT_LM22
LM22 signature matrix  
```{r}
CBST_LM22 = read.table("../_reference/LM22.txt", sep = "\t", header = TRUE)
head(CBST_LM22)
CBST_SFgenes_LM22 = CBST_LM22[as.character(CBST_LM22$Gene.symbol)%in%gene.names,]
CBST.genes = as.character(CBST_SFgenes_LM22$Gene.symbol) # 538 out of 547 
CBST.largeExpression.genes = CBST.genes[CBST.genes %in% large.express.genes.TNK] # 221 out of 538
CBST_SFLEgenes_LM22 = CBST_LM22[as.character(CBST_LM22$Gene.symbol)%in%CBST.largeExpression.genes,]
rm(CBST_SFgenes_LM22)
```

Calculate the (median) expression data of each gene in each of the 3 cell clusters from the k-means analysis
```{r}
# TPM data of signature genes for only TNK cells 
TNKcells_CBSTLEgenes_tpm = TNKcells_tpm[rownames(TNKcells_tpm) %in% CBST.largeExpression.genes,]
TNKcells_CBSTLEgenes_tpm = as.data.table(t(TNKcells_CBSTLEgenes_tpm))
rownames(TNKcells_CBSTLEgenes_tpm) = row.names(df_top50PCs_TNKcells)
colnames(TNKcells_CBSTLEgenes_tpm) = CBST.largeExpression.genes
# Get the median expression level of each gene in each cell cluster
TNKcells_CBSTLEgenes_tpm$grouping = factor(df_top50PCs_TNKcells$KM_3) # cell groupings from the k-means analysis 
TNKcells_CBSTLEgene_expression = TNKcells_CBSTLEgenes_tpm[, lapply(.SD, median), by = grouping] # median expression level of each gene in each cell cluster
dim(TNKcells_CBSTLEgene_expression) # 3 x 222
TNKcells_CBSTLEgene_expression[,1:6]
# Format the gene expression data for each cluster to calculate correlation
compare_expression = data.frame(G1 = as.numeric(TNKcells_CBSTLEgene_expression[1,-1]), G2 = as.numeric(TNKcells_CBSTLEgene_expression[3,-1]), G3 = as.numeric(TNKcells_CBSTLEgene_expression[2,-1]))
row.names(compare_expression) = CBST.largeExpression.genes
head(compare_expression)
```

Check the scatter plot between the three clusters vs. CD4 memory activated T cells, follicular.helper T cells, regulatory T cells, resting NK cells, activated NK cells, and activated Mast cells. On the y-axis is the gene expression in the LM22 matrix for each cell type, and on the x-axis is the gene expression data we calculated from Sade Feldman et al. (2018). 
```{r fig.height=8, fig.width=8}
par(mgp=c(2,0.5,0))
layout(matrix(1:18,nrow=6,ncol = 3,byrow = TRUE))
clusters = c("G1","G2","G3")
cells = c("CD4 memory activated T", "follicular helper T", "regulatory T", "resting NK cells" , "activated NK cells","activated Mast cells")
iter=1
for (i in c(8,9,10,12,13,21)) {
  for (j in 1:3) {
    if (j==1){
      par(mar=c(1.5,3.5,1.5,1))
    } else {
      par(mar=c(1.5,1,1.5,1))
    }
    plot(as.numeric(compare_expression[,j]), as.numeric(CBST_SFLEgenes_LM22[,i]), main = paste0(clusters[j]), ylab = paste0(cells[iter]), xlab=NULL)
  }
  iter = iter+1
}
```

calculate and visualize correlation  
```{r}
iter=1
cormat = matrix(0,nrow = 6,ncol = 3)
colnames(cormat)= c("G1","G2","G3")
rownames(cormat) = c("CD4 memory activated T", "follicular helper T", "regulatory T", "resting NK cells" , "activated NK cells","activated Mast cells")
for (i in c(8,9,10,12,13,21)) {
  for (j in 1:3) {
   cormat[iter,j] = cor(as.numeric(compare_expression[,j]), as.numeric(CBST_SFLEgenes_LM22[,i]),method = "spearman")
  }
  iter = iter+1
}
cormat
```

```{r}
ggcorrplot(t(cormat))
```

Based on the scatter plot, it seems that there are some genes with outlier expression levels in each group (I choosed genes whose expression levels go beyond 10000): 
```{r}
outlier.genes = as.character(CBST_SFLEgenes_LM22$Gene.symbol)[CBST_SFLEgenes_LM22$T.cells.CD4.memory.activated>10000] 
outlier.genes = c(outlier.genes,as.character(CBST_SFLEgenes_LM22$Gene.symbol)[CBST_SFLEgenes_LM22$T.cells.follicular.helper>10000])
outlier.genes = c(outlier.genes,as.character(CBST_SFLEgenes_LM22$Gene.symbol)[CBST_SFLEgenes_LM22$T.cells.regulatory..Tregs.>10000])
outlier.genes = c(outlier.genes,as.character(CBST_SFLEgenes_LM22$Gene.symbol)[CBST_SFLEgenes_LM22$NK.cells.resting > 10000])
outlier.genes = c(outlier.genes,as.character(CBST_SFLEgenes_LM22$Gene.symbol)[CBST_SFLEgenes_LM22$NK.cells.activated > 10000])
outlier.genes = c(outlier.genes,as.character(CBST_SFLEgenes_LM22$Gene.symbol)[CBST_SFLEgenes_LM22$Mast.cells.activated > 10000])
table(outlier.genes)
```

Scatterplot after removing the outliers. On the y-axis is the gene expression in the LM22 matrix for each cell type, and on the x-axis is the gene expression data we calculated from Sade Feldman et al. (2018). 
```{r fig.height=8, fig.width=8}
par(mgp=c(2,0.5,0))
layout(matrix(1:18,nrow=6,ncol = 3,byrow = TRUE))
clusters = c("G1","G2","G3")
cells = c("CD4 memory activated T", "follicular helper T", "regulatory T", "resting NK cells" , "activated NK cells","activated Mast cells")
outlierval = c(10000,10000,10000,10000,10000,10000)
iter=1
for (i in c(8,9,10,12,13,21)) {
  for (j in 1:3) {
    if (j==1){
      par(mar=c(1.5,3.5,1.5,1))
    } else {
      par(mar=c(1.5,1,1.5,1))
    }
    plot(as.numeric(compare_expression[,j])[CBST_SFLEgenes_LM22[,i] <=outlierval[iter]], as.numeric(CBST_SFLEgenes_LM22[,i])[CBST_SFLEgenes_LM22[,i] <=outlierval[iter]], main = paste0(clusters[j]), ylab = paste0(cells[iter]), xlab=NULL)
  }
  iter = iter+1
}
```

Correlation after removing outliers
```{r}
outlierval = c(10000,10000,10000,10000,10000,10000)
iter=1
cormat_rmoutliers = matrix(0,nrow = 6,ncol = 3)
colnames(cormat_rmoutliers)= c("G1","G2","G3")
rownames(cormat_rmoutliers) = c("CD4 memory activated T", "follicular helper T", "regulatory T", "resting NK cells" , "activated NK cells","activated Mast cells")
for (i in c(8,9,10,12,13,21)) {
  for (j in 1:3) {
   cormat_rmoutliers[iter,j] = cor(as.numeric(compare_expression[,j])[CBST_SFLEgenes_LM22[,i] <=outlierval[iter]], as.numeric(CBST_SFLEgenes_LM22[,i])[CBST_SFLEgenes_LM22[,i] <=outlierval[iter]],method = "spearman")
  }
  iter = iter+1
}
cormat_rmoutliers
```

```{r}
ggcorrplot(t(cormat_rmoutliers))
```

```{r}
rm(df_top50PCs_TNKcells)
rm(top50pcs_TNKcells)
rm(TNKcells_tpm)
rm(TNK.cells)
rm(large.express.genes.TNK)
rm(genes.per.cell)
rm(CD8.NK.cells)
rm(reads.per.cell)
```

# Compare the correlation of cell types

Now we have 11 clusters cell types. 
- Compare the expression of LM22 vs. these 11 cell types, by a ggcorrplot, using spearman correlation. 
- Also check the correlation matrix of our 11 cell types. 

```{r}
cell_type = read.table("cell_type.txt",header = TRUE)
table(cell_type$type)
clustered_cells = cell_type$cell[cell_type$type != 0]
```

```{r}
CBST.largeExpression.genes.all = CBST.genes[CBST.genes %in% large.express.genes.all] # 323 genes
CBST_SFLEgenes_LM22 = CBST_LM22[as.character(CBST_LM22$Gene.symbol)%in%CBST.largeExpression.genes.all,]
```

Calculate the (median) expression data of each gene in each of the 3 cell clusters from the k-means analysis
```{r}
# TPM data of signature genes for only TNK cells 
CBSTLEgenes_tpm = sf_tpm[rownames(sf_tpm) %in% CBST.largeExpression.genes.all, ..clustered_cells]
CBSTLEgenes_tpm = as.data.table(t(CBSTLEgenes_tpm)) 
rownames(CBSTLEgenes_tpm) = clustered_cells
colnames(CBSTLEgenes_tpm) = CBST.largeExpression.genes.all
# Get the median expression level of each gene in each of the 11 cell clusters
CBSTLEgenes_tpm$grouping = factor(cell_type$type[match(row.names(CBSTLEgenes_tpm), cell_type$cell)]) 
CBSTLEgene_expression = CBSTLEgenes_tpm[, lapply(.SD, median), by = grouping] # median expression level of each gene in each cell cluster
# dim(CBSTLEgene_expression) # 11 x 324
CBSTLEgene_expression[,1:6]
# Format the gene expression data for each cluster to calculate correlation
compare_expression = data.frame(G1 = as.numeric(CBSTLEgene_expression[11,-1]), 
                                G2 = as.numeric(CBSTLEgene_expression[10,-1]), 
                                G3 = as.numeric(CBSTLEgene_expression[8,-1]),
                                G4 = as.numeric(CBSTLEgene_expression[6,-1]), 
                                G5 = as.numeric(CBSTLEgene_expression[2,-1]),
                                G6 = as.numeric(CBSTLEgene_expression[7,-1]),
                                G7 = as.numeric(CBSTLEgene_expression[5,-1]),
                                G8 = as.numeric(CBSTLEgene_expression[4,-1]),
                                G9 = as.numeric(CBSTLEgene_expression[9,-1]),
                                G10 = as.numeric(CBSTLEgene_expression[1,-1]),
                                G11 = as.numeric(CBSTLEgene_expression[3,-1]))
row.names(compare_expression) = CBST.largeExpression.genes.all
head(compare_expression)
```

calculate and visualize correlation  
```{r}
cormat = matrix(0,nrow = 22, ncol = 11)
colnames(cormat)= colnames(compare_expression)
rownames(cormat) = colnames(CBST_SFLEgenes_LM22)[-1]
for (i in 1:22) {
  for (j in 1:11) {
   cormat[i,j] = cor(as.numeric(compare_expression[,j]), as.numeric(CBST_SFLEgenes_LM22[,i]),method = "spearman")
  }
}
```

```{r}
ggcorrplot(t(cormat))
```

Cell type G5-G11 are strongly correlated with each other. 
- G5: CD4 T regulatory cells
- G6: CD4 T memory cells
- G7: 2482 CD8_B cells
- G8: 3043 CD8_G cells
- G9: 463 NK1 cells
- G10: 1903 NK2 cells
- G11: 1147 NK3 cells
```{r}
cor(compare_expression, method = "spearman")
ggcorrplot(cor(compare_expression, method = "spearman"),type = "upper")
```


# Conclusion

- G1 B cells, G2 Plasma cells, G3 Monocytes/Macrophages, G4 Dendric cells, G7 Regulatory T cells and G10 Memory T-cells could be clearly identified using k-means clustering based on the PCA of high variance genes. The tricky part is to distinguish NK cells and CD8 cells. 
- Recluster G5 - G11 cells didn't really improved the result, and G7 and G10 could also be roughly identified when clustering on all cells, so we changed to recluster cells in G5, G6, G8, G9 and G11. 
- Recluster G5, G6, G8, G9, G11 still didn't give a clear clustering for NK cells, so we moved to excluding all CD8 T cells based on the annotation of the paper, and then cluster the remaining cells.
- There were no strong correlations between the signature gene expressions in LM22 and the gene expression data for each of my NK cell clusters. I tried to remove some outliers, but the result didn't seem to be much different. When using Spearman correlation, regulatory CD4 T cells is the most correlated with the cell clusters. G3 is also slighted correlated with follicular helper CD4 T cells. G1 is slightly correlated with resting NK cells. Finally, we decide to label these cells as NK1, NK2, and NK3 cells for now.

```{r}
gc()
sessionInfo()
```

