---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(Rtsne)
library(R.utils)
library(ggplot2)
```

# Processing data

To make the data easier to work with, I removed the second line containing the patient lables in the first part of the data and write the data to a new data file named GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz. In the following analysis, I directly used this modified dataset. 

```{r eval=FALSE}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_1.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)

sf_tpm1 = sf_tpm1[-1,]
sf_tpm1$V16293 = NULL
gz = gzfile("../scRNAseq/Sade_Feldman_2018/GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz", "w")
write.table(sf_tpm1, gz, quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
close(gz)
rm(gz)
```

```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)
sum(is.na(sf_tpm1))

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))
sum(is.na(sf_tpm2))

ls_tpm = list(sf_tpm1,sf_tpm2)
sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
rm(sf_tpm1)
rm(sf_tpm2)
```

```{r}
dim(sf_tpm) # 55737 x 16292
cell.names = colnames(sf_tpm)[-1]
gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

# Initial QC

## How many reads per cell
```{r}
grouping = rep(1:82, each=200)[1:length(cell.names)]
reads.per.cell = NULL
for (i in 1:82) {
  ss = cell.names[grouping == i]
  temp.reads = unlist(sf_tpm[, lapply(.SD, sum), .SDcols = ss])
  reads.per.cell = c(reads.per.cell,temp.reads)
}
hist(reads.per.cell, main = "Distribution of the amount of transcripts per cell", xlab = "reads per cell")
```

## How many genes are expressed per cell
```{r}
genes.per.cell = NULL
for (i in 1:82) {
  ss = cell.names[grouping == i]
  temp.counts = unlist(sf_tpm[, lapply(.SD, FUN = function(x) sum(x>0)), .SDcols = ss])
  genes.per.cell = c(genes.per.cell,temp.counts)
}
hist(genes.per.cell, main = "Distribution of the number of expressed genes per cell", xlab = "number of espressed genes per cell")
rm(grouping)
```

## How many cells each gene is expressed  
```{r}
gene.grouping = rep(1:28, each = 2000)[1:dim(sf_tpm)[1]]
cells.per.gene = NULL
for (i in 1:28) {
  temp.counts = apply(sf_tpm[gene.grouping==i,], 1, FUN = function(x) sum(x>0))
  cells.per.gene = c(cells.per.gene,temp.counts)
}
hist(log10(cells.per.gene), main = "Distribution of number of cells each genes expressed", xlab = "log10(number of cells) each gene expressed")
rm(gene.grouping)
```

# Dimension reduction and clustering

## Select top 1000 genes with high expression
```{r}
# 5224 genes have zero expression in all cells
unexpressed.genes.id = which(cells.per.gene==0)
unexpressed.genes = gene.names[cells.per.gene==0]
rm(unexpressed.genes.id)
rm(unexpressed.genes)
high.expression = cells.per.gene[order(cells.per.gene,decreasing = TRUE)[1:1000]]
high.expressed.genes = gene.names[order(cells.per.gene,decreasing = TRUE)[1:1000]]
# tmp of the 100 highest expressed genes for all cells
high.expressed.tpm = sf_tpm[order(cells.per.gene,decreasing = TRUE)[1:1000],]
high.expressed.tpm = t(high.expressed.tpm)
colnames(high.expressed.tpm) = high.expressed.genes
high.expressed.tpm[1:10,1:10]
rm(high.expression)
rm(high.expressed.genes)
```

## Perform PCA on selected genes
```{r}
pca.genes = princomp(high.expressed.tpm, cor = TRUE)
rm(high.expressed.tpm)
```

### screen plot
```{r}
eigvals = pca.genes$sdev^2
pve = eigvals/sum(eigvals)
k = length(eigvals)
plot(1:50,pve[1:50],main = "scree plot (first 50 PCs)",type="b",xlab="i-th PC",ylab="Proportion of Variance Explained")
rm(eigvals)
```

### plot of cumulative proportion of variance explained
```{r}
yvec = cumsum(pve)
plot(yvec, type="l", main = "plot of cumulative proportion of variance", xlab = "i-th Principal Components", ylab = "Cumulative variance",ylim=c(0,1))
abline(h=0.8, lty = 2)
rm(yvec)
rm(pve)
```

## Clustering on top 20 PCs

```{r}
top20pcs = pca.genes$scores[,1:20]
head(top20pcs)
dim(top20pcs)
rm(pca.genes)
```

## Calculate TSNE

```{r}
set.seed(9999)
tsne = Rtsne(top20pcs, pca=FALSE)
df_tsne = as.data.frame(tsne$Y)
names(df_tsne) = paste0("topPC_TSNE",seq(ncol(tsne$Y)))
dim(df_tsne)
head(df_tsne)
rm(tsne)
```

## K-means clustering on top 20 PCs

Prepare dataframe for k-means clustering using top 20 PCs
```{r}
df_topPCs = cbind(top20pcs, df_tsne)
rm(df_tsne)
```

Try Kmeans for 5 to 10 clusters 
```{r}
set.seed(9999)
all_num_clust = c(5:10)
for (num_clust in all_num_clust) {
  #num_clust = 5
  cat(paste0("KM with ",num_clust," clusters.\n"))
  kmeans_out = kmeans(top20pcs, centers = num_clust, iter.max = 1e3, nstart = 50, algorithm = "MacQueen")
  print(kmeans_out[c("betweenss","tot.withinss")])
  km_label = paste0("KM_",num_clust)
  df_topPCs[[km_label]] = as.factor(kmeans_out$cluster)
  pt = ggplot(data=df_topPCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=eval(as.name(paste(km_label)))),size = 0.1)+scale_colour_discrete(name=paste(km_label))
  print(pt)
  rm(kmeans_out)
  rm(pt)
}
```

## Compared to the clustering result from Sade-Feldman et al., 2018

```{r}
xnm = "1-s2.0-S0092867418313941-mmc1.xlsx"
table_s1 = readxl::read_xlsx(paste0("../_reference/Sade_Feldman_2018/", xnm), sheet = "Cluster annotation-Fig1B-C", col_names = TRUE)
df_topPCs$sf_cluster = as.factor(table_s1$`Cluster number`)
cat("Cluster annotation from Sade-Feldman et al., 2018, 11 clusters")
ggplot(data=df_topPCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=sf_cluster),size = 0.1)
cat("Cluster annotation from my Kmeans analysis, 10 clusters")
ggplot(data=df_topPCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=KM_10),size = 0.1)
```

```{r}
gc()
sessionInfo()
```


