---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---
```{r echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
source("https://bioconductor.org/biocLite.R")
library(SingleCellExperiment)
library(scater)
library(scran)
library(limma)
library(data.table)
library(Rtsne)
library(R.utils)
library(ggplot2)
```

# Processing data

```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

foo = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), 
            fill = TRUE, nrows=3, header=TRUE)
dim(foo)
foo[,1:3]
cell.names = colnames(foo)
rm(foo)

# Remove the second row that contains labels, 
# and the data is composed of 55737 genes for 16291 cells. 
# The first column is the gene names

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, drop = 16293, skip=2, col.names = cell.names)

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293,  col.names = cell.names)
rm(cell.names)
```

```{r}
dim(sf_tpm) # 55737 x 16292
sf_tpm[1:5, 1:5]
str(sf_tpm)
cell.names = colnames(sf_tpm)[-1]
gene.names = sf_tpm$V1
grouping = rep(1:82, each=200)[1:length(cell.names)-1]
```

# Initial QC

## How many reads per cell
```{r}
reads.per.cell = NULL
for (i in 1:82) {
  ss = cell.names[grouping == i]
  temp.reads = unlist(sf_tpm[, lapply(.SD, sum), .SDcols = ss])
  reads.per.cell = c(reads.per.cell,temp.reads)
}
hist(reads.per.cell, main = "Distribution of the amount of transcripts per cell", xlab = "transcripts per cell")
```

## How many genes are expressed per cell
```{r}
genes.per.cell = NULL
for (i in 1:82) {
  ss = cell.names[grouping == i]
  temp.counts = unlist(sf_tpm[, lapply(.SD, FUN = function(x) sum(x>0)), .SDcols = ss])
  genes.per.cell = c(genes.per.cell,temp.counts)
}
hist(genes.per.cell, main = "Distribution of the number of expressed genes per cell", xlab = "number of espressed genes per cell")
```

## How many cells each gene is expressed  
```{r}
gene.grouping = rep(1:28, each = 2000)[1:dim(sf_tpm)[1]]
cells.per.gene = NULL
for (i in 1:28) {
  temp.counts = apply(sf_tpm[gene.grouping==i,-1], 1, FUN = function(x) sum(x>0))
  cells.per.gene = c(cells.per.gene,temp.counts)
}
hist(log10(cells.per.gene), main = "Distribution of the number of cells each genes expressed", xlab = "log10(number of cells each gene expressed)")
```

# Dimension reduction and clustering

## Select top 1000 genes with high expression
```{r}
# 5224 genes have zero expression in all cells
unexpressed.genes.id = which(cells.per.gene==0)
unexpressed.genes = gene.names[cells.per.gene==0]
high.expression = cells.per.gene[order(cells.per.gene,decreasing = TRUE)[1:1000]]
high.expressed.genes = gene.names[order(cells.per.gene,decreasing = TRUE)[1:1000]]
# tmp of the 100 highest expressed genes for all cells
high.expressed.tpm = sf_tpm[order(cells.per.gene,decreasing = TRUE)[1:1000],-1]
high.expressed.tpm = t(high.expressed.tpm)
colnames(high.expressed.tpm) = high.expressed.genes
#high.expressed.tpm[1:10,1:10]
```

## Perform PCA on selected genes
```{r}
pca.genes = princomp(high.expressed.tpm, cor = TRUE)
```

### screen plot
```{r}
eigvals <- pca.genes$sdev^2
pve = eigvals/sum(eigvals)
k <- length(eigvals)
plot(1:50,pve[1:50],main = "scree plot (first 50 PCs)",type="b",xlab="i-th PC",ylab="Proportion of Variance Explained")
```

### plot of cumulative proportion of variance explained
```{r}
yvec = cumsum(pve)
plot(yvec, type="l", main = "plot of cumulative proportion of variance", xlab = "i-th Principal Components", ylab = "Cumulative variance",ylim=c(0,1))
abline(h=0.8, lty = 2)
```

## Clustering on top 20 PCs

```{r}
top20pcs = pca.genes$scores[,1:20]
head(top20pcs)
dim(top20pcs)
```

## Calculate TSNE

```{r}
set.seed(9999)
tsne = Rtsne(top20pcs, pca=FALSE)
df_tsne = as.data.frame(tsne$Y)
names(df_tsne) = paste0("topPC_TSNE",seq(ncol(tsne$Y)))
dim(df_tsne)
head(df_tsne)
```

## K-means clustering on top 20 PCs

Prepare dataframe for k-means clustering using top 20 PCs
```{r}
df_topPCs = cbind(top20pcs, df_tsne)
dim(df_topPCs)
head(df_topPCs)
cell.info1 = sapply(strsplit(cell.names,split = '_'), "[", 1)
cell.info2 = sapply(strsplit(cell.names,split = '_'), "[", 2)
cell.info3 = sapply(strsplit(cell.names,split = '_'), "[", 3)
df_topPCs$info1 = cell.info1
df_topPCs$info2 = cell.info2
df_topPCs$info3 = cell.info3
```

Try Kmeans for 5 to 15 clusters 
```{r}
set.seed(9999)
all_num_clust = c(5:15)
for (num_clust in all_num_clust) {
  #num_clust = 5
  cat(paste0("KM with ",num_clust," clusters.\n"))
  kmeans_out = kmeans(top20pcs, centers = num_clust, iter.max = 1e3, nstart = 50, algorithm = "MacQueen")
  print(kmeans_out[c("betweenss","tot.withinss")])
  km_label = paste0("KM_",num_clust)
  df_topPCs$km_label = as.factor(kmeans_out$cluster)
  pt = ggplot(data=df_topPCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=km_label),size = 0.1)
  print(pt)
}
```

# Remaining questions:
* How to choose high-variance genes to reduce dimension? It seems that the code in the tutorial for this part is based on a SingleCellExperiment object, but I didn't figured out how to build a SingleCellExperiment object from our TPM format data.
* How to match each cell to specific cell types? I looked through the supplementary tables but still failed to find out information on the type of each cell. 
* There are some functions in the tutorial that I can't use

```{r}
sessionInfo(); q(save="no")
```

