---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(Rtsne)
library(R.utils)
library(ggplot2)
```

# Processing data

## Single cell sequencing data for all cells
```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
rm(sf_tpm1)
rm(sf_tpm2)
```

```{r}
dim(sf_tpm) # 55737 x 16292
cell.names = colnames(sf_tpm)[-1]
gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

## Table of CD8+ T cells 
```{r}
xnm = "1-s2.0-S0092867418313941-mmc2.xlsx"
table_s2 = readxl::read_xlsx(paste0("../_reference/Sade_Feldman_2018/", xnm), sheet = "Cluster annotation-Fig2A-B", col_names = TRUE)
#length(match(table_s2$`Cell Name`, cell.names))
```

## Data for only CD8+ T cells
```{r}
CD8T = table_s2$`Cell Name`
sf_CD8T_tpm = sf_tpm[ , ..CD8T]
rownames(sf_CD8T_tpm) = gene.names
rm(sf_tpm)
```

## How many cells each gene is expressed  
```{r}
gene.grouping = rep(1:28, each = 2000)[1:dim(sf_CD8T_tpm)[1]]
cells.per.gene = NULL
for (i in 1:28) {
  temp.counts = apply(sf_CD8T_tpm[gene.grouping==i,], 1, FUN = function(x) sum(x>0))
  cells.per.gene = c(cells.per.gene,temp.counts)
}
hist(log10(cells.per.gene), main = "Distribution of number of cells each genes expressed", xlab = "log10(number of cells) each gene expressed")
rm(gene.grouping)
```

# Choose high-variance genes

```{r}
# kept 11662 genes
large.express.genes = gene.names[which(cells.per.gene > 200)]
large.expressed.tpm = sf_CD8T_tpm[match(large.express.genes,rownames(sf_CD8T_tpm)),]
# plot the variance of the expression levels for each gene against its mean expression level
genes.mean = apply(large.expressed.tpm, 1, mean)
genes.variance = apply(large.expressed.tpm, 1, var)
df_mv = data.frame(gene=large.express.genes, mean=genes.mean, var=genes.variance)
rm(large.express.genes)
rm(large.expressed.tpm)
rm(genes.mean)
rm(genes.variance)
ggplot(data = df_mv) + 
  geom_point(aes(x=mean,y=var),size = 0.1)
```

```{r}
# divide the genes into 20 equal-size bins based on their mean expression levels
df_mv$grouping = cut_number(df_mv$mean, n=20)
table(df_mv$grouping)
# in each bin, select the first 50 genes with the highest variance
hv.genes.list = by(df_mv, df_mv$grouping, FUN = function(df) return(df$gene[order(df$var,decreasing = TRUE)][1:50]))
hv.genes = as.character(unlist(hv.genes.list))
rm(hv.genes.list)
hvg_tpm = sf_CD8T_tpm[match(hv.genes,rownames(sf_CD8T_tpm)),]
hvg_tpm = t(hvg_tpm)
colnames(hvg_tpm) = hv.genes
rm(df_mv)
```

## Perform PCA on selected genes

```{r}
pca.hvg = princomp(hvg_tpm, cor = TRUE)
rm(hvg_tpm)
```

### screen plot
```{r}
eigvals = pca.hvg$sdev^2
pve = eigvals/sum(eigvals)
plot(1:50,pve[1:50],main = "scree plot (first 50 PCs)",type="b",xlab="i-th PC",ylab="Proportion of Variance Explained")
rm(eigvals)
```

### plot of cumulative proportion of variance explained
```{r}
yvec = cumsum(pve)
plot(yvec, type="l", main = "plot of cumulative proportion of variance", xlab = "i-th Principal Components", ylab = "Cumulative variance",ylim=c(0,1))
abline(h=0.8, lty = 2)
rm(yvec)
rm(pve)
```

```{r}
pca.scores = pca.hvg$scores
rm(pca.hvg)
```

# Clustering on top 50 PCs

```{r}
top50pcs = pca.scores[,1:50]
rm(pca.scores)
#head(top50pcs)
```

## Calculate TSNE

```{r}
set.seed(9999)
tsne = Rtsne(top50pcs, pca=FALSE)
df_tsne = as.data.frame(tsne$Y)
names(df_tsne) = paste0("topPC_TSNE",seq(ncol(tsne$Y)))
rm(tsne)
```

## K-means clustering on top 50 PCs

Prepare dataframe for k-means clustering using top 50 PCs
```{r}
df_top50PCs = cbind(top50pcs, df_tsne)
rm(df_tsne)
#head(df_top50PCs)
rm(top50pcs)
```

Try Kmeans for 2 to 8 clusters 
```{r}
set.seed(9999)
all_num_clust = c(2:8)
for (num_clust in all_num_clust) {
  cat(paste0("KM with ",num_clust," clusters.\n"))
  kmeans_out = kmeans(df_top50PCs[,1:50], centers = num_clust, iter.max = 1e3, nstart = 50, algorithm = "MacQueen")
  print(kmeans_out[c("betweenss","tot.withinss")])
  km_label = paste0("KM_",num_clust)
  df_top50PCs[[km_label]] = as.factor(kmeans_out$cluster)
  pt = ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=eval(as.name(paste(km_label)))),size = 0.3)+scale_colour_discrete(name=paste(km_label))+guides(color = guide_legend(override.aes = list(size = 2)))
  print(pt)
  rm(kmeans_out)
  rm(pt)
}
```

# Compared to the clustering result from Sade-Feldman et al., 2018

```{r}
# CD8_G cluster and CD8_B cluster label for each CD8+ T cell
df_top50PCs$CD8T = as.factor(table_s2$Cluster)

# 6 CD8+ T cell clusters in Figure 4
xnm = "1-s2.0-S0092867418313941-mmc4.xlsx"
table_s4 = readxl::read_xlsx(paste0("../_reference/Sade_Feldman_2018/", xnm), sheet = "Cluster annotation-Fig4A-B", col_names = TRUE)
df_top50PCs$sf_cluster = as.factor(table_s4$Cluster)
```

## CD8_G cluster and CD8_B cluster annotation from Sade-Feldman et al., 2018
```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=CD8T),size = 0.3)+guides(color = guide_legend(override.aes = list(size = 2)))
```

In the K-means analysis result using 3 clusters, the cluster of CD8_B cells and that of CD8_D cells could be roughly identified. Cluster 1 contains all CD8_B cells, cluster 2 contains mostly CD8_B cells, and cluster 3 contains mostly CD8_G cells. 

```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=KM_3),size = 0.3)+guides(color = guide_legend(override.aes = list(size = 2)))
```

```{r}
table(df_top50PCs$KM_3, df_top50PCs$CD8T)
```

## Fine clustering of CD8+ T cells in Sade-Feldman et al., 2018

```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=sf_cluster),size = 0.3)+guides(color = guide_legend(override.aes = list(size = 2)))
```

K-means analysis using 8 clusters identified the most CD8+ T cells clusters defined by the paper.   
CD8_1 cells are mostly covered by clusters 3 and 5, but part of them are also in cluster 1.  
CD8_2 cells could not be clearly identified by one or two clusters.  
CD8_3 cells are mostly covered by cluster 1, but some of them could not be identified from CD8_1 cells and CD8_2 cells in cluster 6.  
CD8_4 cells are mostly covered by cluster 7, but some of them are mixed in cluster 4.  
CD8_5 cells could not be clearly identified from CD8_3 cells, CD8_4 cells, and CD8_6 cells cluster 4.  
Cluster 2 contains roughly a half of CD8_6 cells, the other half is mixed in clusters 1, 4, 8.  

```{r}
ggplot(data=df_top50PCs)+geom_point(aes(x=topPC_TSNE1,y=topPC_TSNE2,color=KM_8),size = 0.3)+guides(color = guide_legend(override.aes = list(size = 2)))
```

```{r}
table(df_top50PCs$KM_8, df_top50PCs$sf_cluster)
```

```{r}
rm(df_top50PCs)
```

```{r}
gc()
sessionInfo()
```
