---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library("MAST")
library(data.table)
library(R.utils)
```

# Processing data

```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
rm(sf_tpm1)
rm(sf_tpm2)

sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
```

```{r}
#dim(sf_tpm) # 55737 x 16292
#cell.names = colnames(sf_tpm)[-1]
gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

```{r}
cell_type = read.table("cell_type.txt",header = TRUE)
cell_type$cell = as.character(cell_type$cell)
table(cell_type$type)
clustered_cells = cell_type$cell[cell_type$type != 0]
cluster_sf_tpm = sf_tpm[, ..clustered_cells]
rm(sf_tpm)
rm(clustered_cells)
```

# Average expression of each gene in each cell type

First calculate the expression of each gene across all cells

```{r}
all.expression = apply(cluster_sf_tpm, 1, sum)
```

u_1 - u_8: Average expression of each gene for each cell type
o_1 - o_8: For a particular cell type, the average expression of each gene in all other cells

```{r}
average_expression = data.frame(gene = gene.names)

for (i in 1:8) {
  cc = cell_type$cell[cell_type$type == i] # select cells in each cell cluster
  cluster.exp = apply(cluster_sf_tpm[, ..cc], 1, sum) # total expression of each gene across all cells in a particular cluster
  exclu.exp = all.expression - cluster.exp # total expression of each gene across all other cells
  aveg.cluster.exp = cluster.exp / length(cc) # divide by the number of cells in the cluster to get the average expression
  aveg.exclu.exp = exclu.exp / (ncol(cluster_sf_tpm) - length(cc)) # divide by the number of all other cells to get the average expression
  fc = aveg.exclu.exp / aveg.cluster.exp
  ulabel = paste0("u_",i)
  average_expression[[ulabel]] = aveg.cluster.exp
  olabel = paste0("o_",i)
  average_expression[[olabel]] = aveg.exclu.exp
  fclabel = paste0("fc_",i)
  average_expression[[fclabel]] = fc
}

rm(cc)
rm(cluster.exp)
rm(exclu.exp)
rm(aveg.cluster.exp)
rm(aveg.exclu.exp)
rm(fc)
```

```{r}
head(average_expression)
```

# Testing differential gene expression

##  Loading and transforming data

The tutorial mentioned that the data should be a matrix of log2 + 1 transcripts per million (TPM). The data from Sade-Feldman et al. (2018) are in TPM format, so I transfromed it with log2 + 1: 

```{r}
# cluster_sf_tpm is a big data table of TPM data, with one row per gene and one column per cell
cluster_sf_tpm_log = log2(1 + as.matrix(cluster_sf_tpm)) #log transformed data

dim(cluster_sf_tpm_log)
cluster_sf_tpm_log[1:10, 1:10]
row.names(cluster_sf_tpm_log) = gene.names

# get the cell names from the data
cell.id = colnames(cluster_sf_tpm_log)   
# gene names are saved in the vector gene.names
```

## Calculating the CDR rate: number of genes detected for each cell
```{r}
cdr = rep(NA, ncol(cluster_sf_tpm))
for (i in 1:8) {
  ss = cell_type$cell[cell_type$type == i]
  temp.counts = unlist(cluster_sf_tpm[, lapply(.SD, FUN = function(x) sum(x>0)), .SDcols = ss])
  cdr[match(ss, cell.id)] = temp.counts
}
rm(ss)
rm(temp.counts)
rm(cluster_sf_tpm)
```

## cell type label: diagnosis

```{r}
# cell type label
cell_group = factor(cell_type$type[cell_type$type != 0])
rm(cell_type)
```

## Preparing FromMatrix

```{r eval=FALSE}
fData = data.frame(primerid = gene.names)
cData = data.frame(wellKey = cell.id)
#nrow(cluster_sf_tpm_log) == nrow(fData)
#ncol(cluster_sf_tpm_log) == nrow(cData)
rm(cell.id)
rm(gene.names)

# Run-out-of-memory error at this step. I'll try to use cluster computing for this. 
sca = FromMatrix(cluster_sf_tpm_log, cData, fData) 
colData(sca)$cngeneson = scale(cdr)
colData(sca)$diagnosis = cell_group

colData(sca)
```

## Testing

```{r eval=FALSE}
getOption("mc.cores")

date()
b0 = zlm(formula = ~ diagnosis + cngeneson, sca = sca, parallel = TRUE)
date()

lrt0 = lrTest(b0, "diagnosis")

dim(lrt0)
lrt0[1,,]
mast_pval_glm = apply(lrt0, 1, function(x){x[3,3]})
```




