---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library("MAST")
library(data.table)
library(R.utils)
```

# Processing data

```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
rm(sf_tpm1)
rm(sf_tpm2)

sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
```

```{r}
#dim(sf_tpm) # 55737 x 16292
#cell.names = colnames(sf_tpm)[-1]
gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

```{r}
cell_type = read.table("cell_type.txt",header = TRUE)
cell_type$cell = as.character(cell_type$cell)
table(cell_type$type)
clustered_cells = cell_type$cell[cell_type$type != 0]
cluster_sf_tpm = sf_tpm[, ..clustered_cells] # 55737 x 11264
rm(sf_tpm)
rm(clustered_cells)
```

# Average expression of each gene in each cell type

First calculate the expression of each gene across all cells

```{r}
all.expression = apply(cluster_sf_tpm, 1, sum)
```

u_1 - u_8: Average expression of each gene for each cell type
o_1 - o_8: For a particular cell type, the average expression of each gene in all other cells

```{r}
average_expression = data.frame(gene = gene.names)

for (i in 1:8) {
  cc = cell_type$cell[cell_type$type == i] # select cells in each cell cluster
  cluster.exp = apply(cluster_sf_tpm[, ..cc], 1, sum) # total expression of each gene across all cells in a particular cluster
  exclu.exp = all.expression - cluster.exp # total expression of each gene across all other cells
  aveg.cluster.exp = cluster.exp / length(cc) # divide by the number of cells in the cluster to get the average expression
  aveg.exclu.exp = exclu.exp / (ncol(cluster_sf_tpm) - length(cc)) # divide by the number of all other cells to get the average expression
  fc = aveg.exclu.exp / aveg.cluster.exp
  ulabel = paste0("u_",i)
  average_expression[[ulabel]] = aveg.cluster.exp
  olabel = paste0("o_",i)
  average_expression[[olabel]] = aveg.exclu.exp
  fclabel = paste0("fc_",i)
  average_expression[[fclabel]] = fc
}

rm(cc)
rm(cluster.exp)
rm(exclu.exp)
rm(aveg.cluster.exp)
rm(aveg.exclu.exp)
rm(fc)
```

```{r}
head(average_expression)
```

# Testing differential gene expression

##  Loading and transforming data

The tutorial mentioned that the data should be a matrix of log2 + 1 transcripts per million (TPM). The data from Sade-Feldman et al. (2018) are in TPM format, so I transfromed it with log2 + 1: 

```{r}
# cluster_sf_tpm is a big data table of TPM data, with one row per gene and one column per cell
cluster_sf_tpm_log = log2(1 + as.matrix(cluster_sf_tpm)) #log transformed data

dim(cluster_sf_tpm_log)
cluster_sf_tpm_log[1:10, 1:10]
row.names(cluster_sf_tpm_log) = gene.names

# get the cell names from the data
cell.id = colnames(cluster_sf_tpm_log)   
# gene names are saved in the vector gene.names
```

## Calculating the CDR rate: number of genes detected for each cell
```{r}
cdr = rep(NA, ncol(cluster_sf_tpm))
for (i in 1:8) {
  ss = cell_type$cell[cell_type$type == i]
  temp.counts = unlist(cluster_sf_tpm[, lapply(.SD, FUN = function(x) sum(x>0)), .SDcols = ss])
  cdr[match(ss, cell.id)] = temp.counts
}
rm(ss)
rm(temp.counts)
rm(cluster_sf_tpm)
```

## Preparing FromMatrix

```{r eval=FALSE}
fData = data.frame(primerid = gene.names)
cData = data.frame(wellKey = cell.id)
#nrow(cluster_sf_tpm_log) == nrow(fData)
#ncol(cluster_sf_tpm_log) == nrow(cData)

# Run-out-of-memory error at this step. I'll try to use cluster computing for this. 
sca = FromMatrix(cluster_sf_tpm_log, cData, fData) 
# On bayes.biostat.washington.edu:
# class: SingleCellAssay 
# dim: 55737 11264 
# metadata(0):
# assays(1): et
# rownames(55737): TSPAN6 TNMD ... AP000230.1 RP11-80H18.4
# rowData names(1): primerid
# colnames(11264): A11_P4_M11 A2_P4_M11 ... B8_P7_M74_T_enriched
#   A12_P6_MMD8_L001_T_enriched
# colData names(1): wellKey
# reducedDimNames(0):
# spikeNames(0):

colData(sca)$cngeneson = scale(cdr)
for(i in 1:8){
  label = paste0("G_",i)
  colData(sca)[[label]] = "control"
  caseid = match(cell_type$cell[cell_type$type == i],colData(sca)$wellKey)
  colData(sca)[[label]][caseid] = "case"
  colData(sca)[[label]] = as.factor(colData(sca)[[label]])
}

colData(sca)
```

DataFrame with 11264 rows and 10 columns
                                                wellKey          cngeneson
                                            <character>           <matrix>
A11_P4_M11                                   A11_P4_M11  0.472000836313483
A2_P4_M11                                     A2_P4_M11 -0.201989553675478
A3_P3_M11                                     A3_P3_M11   0.60842789410581
A4_P3_M11                                     A4_P3_M11  0.746891176641307
A4_P4_M11                                     A4_P4_M11 0.0209770557015344
...                                                 ...                ...
C8_P12_MMD2-36B_T_enriched   C8_P12_MMD2-36B_T_enriched -0.424956163052491
D5_P12_MMD2-36B_T_enriched   D5_P12_MMD2-36B_T_enriched -0.529821737325698
F3_P12_MMD2-36B_T_enriched   F3_P12_MMD2-36B_T_enriched -0.169409957784773
B8_P7_M74_T_enriched               B8_P7_M74_T_enriched -0.730389874527851
A12_P6_MMD8_L001_T_enriched A12_P6_MMD8_L001_T_enriched    0.3589903630676
                                 G_1      G_2      G_3      G_4      G_5
                            <factor> <factor> <factor> <factor> <factor>
A11_P4_M11                   control  control  control  control     case
A2_P4_M11                    control  control  control  control  control
A3_P3_M11                    control  control  control  control     case
A4_P3_M11                    control  control  control  control  control
A4_P4_M11                    control  control  control  control  control
...                              ...      ...      ...      ...      ...
C8_P12_MMD2-36B_T_enriched   control  control  control  control  control
D5_P12_MMD2-36B_T_enriched   control  control  control  control  control
F3_P12_MMD2-36B_T_enriched   control  control  control  control  control
B8_P7_M74_T_enriched         control  control  control  control  control
A12_P6_MMD8_L001_T_enriched  control  control  control  control     case
                                 G_6      G_7      G_8
                            <factor> <factor> <factor>
A11_P4_M11                   control  control  control
A2_P4_M11                    control  control     case
A3_P3_M11                    control  control  control
A4_P3_M11                    control     case  control
A4_P4_M11                    control     case  control
...                              ...      ...      ...
C8_P12_MMD2-36B_T_enriched      case  control  control
D5_P12_MMD2-36B_T_enriched      case  control  control
F3_P12_MMD2-36B_T_enriched      case  control  control
B8_P7_M74_T_enriched            case  control  control
A12_P6_MMD8_L001_T_enriched  control  control  control

## Testing

```{r eval=FALSE}
# detectCores() # 20
# options(mc.cores=10) # Got failuresError in mcfork() : unable to fork, possible reason: Cannot allocate memory

zlmlist = list()
# for (i in 1:8) {
#   label = paste0("G_",i)
#   zlmlist[label] = zlm(formula = ~ eval(as.name(paste(label))) + cngeneson, sca = sca, parallel=TRUE)
# }
# approximately one hour for each run
zlmlist["G_1"] = zlm(formula = ~ G_1 + cngeneson, sca = sca, parallel = TRUE) 
zlmlist["G_2"] = zlm(formula = ~ G_2 + cngeneson, sca = sca, parallel = TRUE)
zlmlist["G_3"] = zlm(formula = ~ G_3 + cngeneson, sca = sca, parallel = TRUE)
zlmlist["G_4"] = zlm(formula = ~ G_4 + cngeneson, sca = sca, parallel = TRUE)
zlmlist["G_5"] = zlm(formula = ~ G_5 + cngeneson, sca = sca, parallel = TRUE)
zlmlist["G_6"] = zlm(formula = ~ G_6 + cngeneson, sca = sca, parallel = TRUE)
zlmlist["G_7"] = zlm(formula = ~ G_7 + cngeneson, sca = sca, parallel = TRUE)
zlmlist["G_8"] = zlm(formula = ~ G_8 + cngeneson, sca = sca, parallel = TRUE)

# zlmlist
# $G_1
# Fitted zlm on 55737 genes and 11264 cells.
#  Using BayesGLMlike ~ G_1 + cngeneson 
# 
# $G_2
# Fitted zlm on 55737 genes and 11264 cells.
#  Using BayesGLMlike ~ G_2 + cngeneson 
# 
# $G_3
# Fitted zlm on 55737 genes and 11264 cells.
#  Using BayesGLMlike ~ G_3 + cngeneson 
# 
# $G_4
# Fitted zlm on 55737 genes and 11264 cells.
#  Using BayesGLMlike ~ G_4 + cngeneson 
# 
# $G_5
# Fitted zlm on 55737 genes and 11264 cells.
#  Using BayesGLMlike ~ G_5 + cngeneson 
# 
# $G_6
# Fitted zlm on 55737 genes and 11264 cells.
#  Using BayesGLMlike ~ G_6 + cngeneson 
# 
# $G_7
# Fitted zlm on 55737 genes and 11264 cells.
#  Using BayesGLMlike ~ G_7 + cngeneson 
# 
# $G_8
# Fitted zlm on 55737 genes and 11264 cells.
#  Using BayesGLMlike ~ G_8 + cngeneson

lrtlist = list()
# running
for (i in 1:8) {
  lrtlist[paste0("G_",i)] = lrTest(zlmlist[[i]], paste0("G_",i))
}

dim(zlmlist[1]) 
zlmlist[[1]]

mast_pval_glm_list = lapply(zlmlist, FUN = function(lrt0) apply(lrt0, 1, function(x){x[3,3]}))
```




