---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(Rtsne)
library(R.utils)
library(ggplot2)
library(ggcorrplot)
```

# Processing data

## Sade-Feldman(2018) data
```{r}
fnm1 = "GSE120575_Sade_Feldman_melanoma_single_cells_nolabel_1.txt.gz"
fnm2 = "GSE120575_Sade_Feldman_melanoma_single_cells_2.txt.gz"

sf_tpm1 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm1), fill = TRUE, header = TRUE)

sf_tpm2 = fread(paste0("../scRNAseq/Sade_Feldman_2018/", fnm2), fill = TRUE, drop = 16293, col.names = colnames(sf_tpm1))

ls_tpm = list(sf_tpm1,sf_tpm2)
sf_tpm = rbindlist(ls_tpm)

rm(ls_tpm)
rm(sf_tpm1)
rm(sf_tpm2)
```

```{r}
dim(sf_tpm) # 55737 x 16292
gene.names = sf_tpm$V1
rownames(sf_tpm) = gene.names
sf_tpm$V1 = NULL
```

number of cells that each gene is expressed in, to select genes that have expression in a certain amount of cells 
```{r}
gene.grouping = rep(1:28, each = 2000)[1:dim(sf_tpm)[1]]
cells.per.gene = NULL
for (i in 1:28) {
  temp.counts = rowSums(sf_tpm[gene.grouping==i,] > 0)
  cells.per.gene = c(cells.per.gene,temp.counts)
}
large.express.genes.all = gene.names[which(cells.per.gene > 500)] # kept 12705 genes
rm(gene.grouping)
```

## 11 cell types we defined from Sade-Feldman(2018) data

```{r}
cell_type = read.table("cell_type.txt",header = TRUE,sep = "\t")
table(cell_type$type)
clustered_cells = cell_type$cell[cell_type$type != "unclustered"]
# select TPM data for 12705 genes with large expression in 14777 grouped cells
clustered_tpm_LEgenes = sf_tpm[rownames(sf_tpm) %in% large.express.genes.all, ..clustered_cells] 
rownames(clustered_tpm_LEgenes) = large.express.genes.all
rm(sf_tpm)
```

## percentages of zero expression genes for each cell in each cell type
```{r}
# calculate the percentages of genes with zero expression for each cell
zeros.per.cell = colSums(clustered_tpm_LEgenes==0) / nrow(clustered_tpm_LEgenes)
zero.percentages = data.frame(percentage.of.zero = zeros.per.cell, grouping = factor(cell_type$type[match(clustered_cells, cell_type$cell)]))
head(zero.percentages)
# distribution of the percentages of zero expression genes for cells by cell type
ggplot(data = zero.percentages) + 
  geom_histogram(aes(x=percentage.of.zero), binwidth = 0.01) + 
  facet_wrap(~grouping) + 
  ggtitle("Distribution of the percentages of zero expression genes")
rm(zeros.per.cell)
```

Based on the graph, the spread of distribution of the percentages of zero expression genes is quite different for each cell type, so I didn't use a single number as the cutoff to choose which cells to keep, instead, I select 100 cells with the smallest percentages of zero expression genes.
```{r}
fewzero_cells = unname(unlist(by(zero.percentages, zero.percentages$grouping, FUN = function(x) rownames(x)[order(x$percentage.of.zero)][1:100])))
```

## LM22 signature matrix  
- select genes that appear both in the LM22 signature matrix and the SF data
- in those genes, further select genes that have expression in a relatively large amount of cells
```{r}
CBST_LM22 = read.table("../_reference/LM22.txt", sep = "\t", header = TRUE, as.is = TRUE)
head(CBST_LM22)
CBST.SFgenes = CBST_LM22$Gene.symbol[CBST_LM22$Gene.symbol%in%gene.names] # 538 out of 547 
CBST.largeExpression.genes = CBST.SFgenes[CBST.SFgenes %in% large.express.genes.all] # 323 out of 538
CBST_SFLEgenes_LM22 = CBST_LM22[CBST_LM22$Gene.symbol%in%CBST.largeExpression.genes,] # 323 x 23
head(CBST_SFLEgenes_LM22)
```

# Compare the correlation of cell types
- Compare the correlation between our 11 cell types. 
- Compare the correlation between the cell types in the LM22 vs. the 11 cell types we had. 
- Two types of correlation: using all cells for each cell type to calculate mean gene expression, or calculate the mean expression using 100 cells with smallest percentages of zero expression genes. 

## Correlations between our 11 cell types
### using mean expression of large expression genes shared between SF data and LM22 matrix
```{r}
LEgenes_CBST_tpm = as.data.table(t(clustered_tpm_LEgenes[large.express.genes.all %in% CBST.largeExpression.genes,]))
rownames(LEgenes_CBST_tpm) = clustered_cells
colnames(LEgenes_CBST_tpm) = CBST.largeExpression.genes
LEgenes_CBST_tpm[1:5,1:5]
LEgenes_CBST_tpm$grouping = factor(cell_type$type[match(row.names(LEgenes_CBST_tpm), cell_type$cell)])
table(LEgenes_CBST_tpm$grouping)
LEgene_CBST_expression = LEgenes_CBST_tpm[, lapply(.SD, mean), by = grouping] # 11 x 324
LEgene_CBST_expression = LEgene_CBST_expression[order(LEgene_CBST_expression$grouping),]
LEgene_CBST_expression[,1:6]
LEgene_CBST_compare_expression = as.data.frame(t(LEgene_CBST_expression[,-1]))
colnames(LEgene_CBST_compare_expression) = LEgene_CBST_expression$grouping
head(LEgene_CBST_compare_expression)
rm(LEgene_CBST_expression)
rm(LEgenes_CBST_tpm)
```

The patterns of the correlation roughly match our expectation: 
- Plasma cells, Monocytes/Macrophages, Dendritic cells and B cells do no have major correlation with other cell types, matches the clustering result that these 4 cell types form most distinct clusters
- The 3 NK cell clusters are closely related with each other
- CD8 cells, NK cells, and CD4 T cells are closely related
```{r}
cormat_LE.CBSTgenes_11type = cor(LEgene_CBST_compare_expression, method = "spearman")
ggcorrplot(cormat_LE.CBSTgenes_11type,type = "upper")
```

### using mean expression for large expression genes shared between SF data and LM22 matrix over cells with fewer zero gene expression 
```{r}
LEgenes_CBST_few0_tpm = as.data.table(t(clustered_tpm_LEgenes[large.express.genes.all %in% CBST.largeExpression.genes, ..fewzero_cells])) 
rownames(LEgenes_CBST_few0_tpm) = fewzero_cells
colnames(LEgenes_CBST_few0_tpm) = CBST.largeExpression.genes
LEgenes_CBST_few0_tpm[1:5,1:5]
LEgenes_CBST_few0_tpm$grouping = factor(cell_type$type[match(row.names(LEgenes_CBST_few0_tpm), cell_type$cell)])
table(LEgenes_CBST_few0_tpm$grouping)

LEgene_CBST_few0_expression = LEgenes_CBST_few0_tpm[, lapply(.SD, mean), by = grouping] # 11 x 324
LEgene_CBST_few0_expression[,1:6]
LEgene_CBST_few0_compare_expression = as.data.frame(t(LEgene_CBST_few0_expression[,-1]))
colnames(LEgene_CBST_few0_compare_expression) = LEgene_CBST_few0_expression$grouping
head(LEgene_CBST_few0_compare_expression)
rm(LEgene_CBST_few0_expression)
rm(LEgenes_CBST_few0_tpm)
```

The result using cells with fewer zero expression genes is not very different from the result using all cells. 
```{r}
cormat_LE.CBST.few0genes_11type = cor(LEgene_CBST_few0_compare_expression, method = "spearman")
ggcorrplot(cormat_LE.CBST.few0genes_11type,type = "upper")
```

## Correlations between our 11 cell types and the cell types in the LM22 matrix
### using mean expression of large expression genes shared between SF data and LM22 matrix
```{r}
cormat_LM22_11type = matrix(0,nrow = 22, ncol = 11)
colnames(cormat_LM22_11type)= colnames(LEgene_CBST_compare_expression)
rownames(cormat_LM22_11type) = colnames(CBST_SFLEgenes_LM22)[-1]
for (i in 1:22) {
  for (j in 1:11) {
   cormat_LM22_11type[i,j] = cor(as.numeric(LEgene_CBST_compare_expression[,j]), as.numeric(CBST_SFLEgenes_LM22[,i+1]),method = "spearman")
  }
}
```

The correlations between our 11 cell types and the cell types in the LM22 matrix still have some unexpected patterns
```{r fig.height=5, fig.width=4}
ggcorrplot(t(cormat_LM22_11type))
```

### using mean expression for large expression genes shared between SF data and LM22 matrix over cells with fewer zero gene expression 
```{r}
cormat_LM22_11type_few0 = matrix(0,nrow = 22, ncol = 11)
colnames(cormat_LM22_11type_few0)= colnames(LEgene_CBST_few0_compare_expression)
rownames(cormat_LM22_11type_few0) = colnames(CBST_SFLEgenes_LM22)[-1]
for (i in 1:22) {
  for (j in 1:11) {
   cormat_LM22_11type_few0[i,j] = cor(as.numeric(LEgene_CBST_few0_compare_expression[,j]), as.numeric(CBST_SFLEgenes_LM22[,i+1]),method = "spearman")
  }
}
```

The result using cells with fewer zero expression genes is not very different from the result using all cells. 
```{r fig.height=5, fig.width=4}
ggcorrplot(t(cormat_LM22_11type_few0))
```




